---
title: "iPSC Motor Neurons and Spinal Cord Cell Refs Data"
author: "Jenny L Smith"
date: "`R Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Set-up

```{r set-up, echo=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80),
                      tidy=TRUE,
                      fig.align='center',
                      fig.width = 10, fig.height = 10,
                      eval = TRUE)

options(stringsAsFactors = FALSE, max.print = 100)
options(Seurat.object.assay.version = "v5")
table = function (..., useNA = 'ifany') base::table(..., useNA = useNA)
```

```{r message=FALSE,warning=FALSE}
library(dplyr)
library(tidyr)
library(magrittr)
library(stringr)
library(tibble)
library(glue)

library(ggplot2)
library(plotly)
library(RColorBrewer)

library(Seurat)
library(SeuratDisk)
library(ontologyIndex)
library(reticulate)
```

```{r}
conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter","dplyr")
conflicted::conflict_prefer("saveRDS","base")
```

<https://github.com/bnprks/BPCells/issues/6>

remotes::install_github("bnprks/BPCells")

# Parallelization

```{r}
library(future)
# check the current active plan
plan()

# plan("multisession", workers = 2)
# plan()
options(future.globals.maxSize = 10000 * 1024^2)
```
# Define Functions


```{r}
clean_standardize_annots <- function(seurat_obj){
  
  age_tissue_clean <- seurat_obj@meta.data %>% 
    mutate(tissue_clean = case_when(
    dataset == "Allen_M1" ~ "adult_motor_cortex",
    dataset == "Rayon_2021" ~  paste0(gsub("human_|_rep[0-9]|spinalcord","", orig.ident), "_spinal_cord"),
    dataset == "yadav_2023" ~ gsub("adult_Lumbar_spinal_cord_", "adult", tolower(tissue))), 
    
  age_clean = case_when(
      donor == "H18.30.002" ~ "50_yrs",
      donor == "H18.30.001" ~ "60_yrs",
      TRUE ~ gsub("\\~", "", age_clean))) %>% 
    select(age_clean, tissue, tissue_clean)
  
  seurat_obj$tissue <- age_tissue_clean$tissue_clean
  seurat_obj$age_clean <- age_tissue_clean$age_clean


  annots_table_clean <- seurat_obj@meta.data %>% 
    mutate(class_label_standardized = case_when(
      grepl("Astr|Oligo|Micro", top_level_annotation) ~ "glia",
      grepl("OPC", top_level_annotation) ~ "glial_progenitor",
      grepl("Endo|Blood|Epe|Ery|Hema|Lymph|Meso|Menin|Schwann|VLMC|Pericytes", top_level_annotation) ~ "non-neuronal",
      grepl("GABAergic|Glutamatergic", class_label) ~ tolower(class_label),
      grepl("Neuro|neuro", top_level_annotation) ~ "neuronal",
      grepl("[Pp]rogenitor", top_level_annotation) ~ "neuronal_progenitor",
      TRUE ~ ""), 
      
      top_level_standardized = case_when(
        grepl("Blood|Ery|Hema", top_level_annotation) ~ "blood",
        grepl("^Mesoderm", top_level_annotation) ~ "mesoderm",
        
        grepl("^Astr", top_level_annotation) ~ "astrocyte",
        grepl("^Endo", top_level_annotation) ~ "endothelial",
        grepl("Ependymal", top_level_annotation) ~ "epidymal",
        grepl("Micro", top_level_annotation) ~ "microglia",
  
        grepl("Oligo", top_level_annotation) ~ "oligodendrocytes",
        # grepl("N[Uu][Ll]", subtype_annotation) ~ "esc_neuron",
        grepl("Neuro", top_level_annotation) & !grepl("Moto|^MN", subtype_annotation) ~ "neuron", 
        # grepl("Neuro", top_level_annotation) & grepl("IEG", subtype_annotation) ~ "IEG_neuron",
        # grepl("Neuro", top_level_annotation) & !grepl("Moto|^MN", subtype_annotation) ~ paste0("sc_",gsub("^.+(dorsal|ventral|mid).+", "\\1", description),"_neuron"),
        grepl("Neuro", top_level_annotation) & grepl("Moto|^MN", subtype_annotation) ~ "motorneurons",
        grepl("DRG_Progenitor", top_level_annotation) ~ "progenitor",
        TRUE ~ tolower(top_level_annotation)
      ), 
      
      subtype_standardized = case_when(
        grepl("pMN", subtype_annotation) ~ "pMN",
        grepl("Neuro", top_level_annotation) & !grepl("Moto|^MN", subtype_annotation) ~ gsub("-[0-9]{1,2}$","", subtype_annotation) %>% 
                                                                                                  gsub(".+\\s(.+)$", "\\1", .) %>% gsub("\\/|-","_", .),
        TRUE ~ top_level_standardized)) %>% 
    select(matches("standardized"))
  
  # head(annots_table_clean)
  dim(annots_table_clean)
  
  
  # Update the metadata
  if(!all(colnames(annots_table_clean) %in% colnames(seurat_obj@meta.data))){
    seurat_obj <- AddMetaData(seurat_obj, annots_table_clean)
  }
  
  return(seurat_obj)
}
```

```{r}
library(plotly)
make_3D_umap <- function(sobj_3d, dims = 1:30, idents_column, selected_colors=NULL){
    "
    idents_column is for labeling. ex) 'RNA_snn_res.0.7'
    "
    # 3D UMAP
    plot.data <- FetchData(object = sobj_3d, vars = c("umap_1", "umap_2", "umap_3", idents_column))
    plot.data$label <- paste(rownames(plot.data))
    
    if(is.null(selected_colors)){
      selected_colors <- c("lightseagreen", "gray50",  "darkgreen", "red4", "red", "turquoise4","black","yellow4","royalblue1","lightcyan3",
                 "peachpuff3","lightskyblue","gray20","orange2","royalblue4","yellow3","gray80","darkorchid1","lawngreen","plum2", 
                 "darkorange4","darkmagenta","mediumorchid1","lightslateblue","midnightblue", "tan2") #26
    }

    
    fig <- plot_ly(data = plot.data, 
        x = ~umap_1, y = ~umap_2, z = ~umap_3, 
        color = ~get(idents_column), 
        colors = selected_colors, 
        type = "scatter3d", 
        mode = "markers", 
        marker = list(size = 3, width=2), # controls size of points
        text=~label, # This is that extra column we made earlier for which we will use for cell ID
        hoverinfo="text") #When you visualize your plotly object, hovering your mouse pointer over a point shows cell names
    
    
    # xaxis
    axx <- list( nticks = 4, range = c(-10,10)) #select range of xaxis
 
    # yaxis
    axy <- list(nticks = 4,  range = c(-10,10)) 
    
    #zaxis
    axz <- list(nticks = 4, range = c(-10,10)) 
    
    fig <- fig %>% layout(scene = list(xaxis=axx,yaxis=axy,zaxis=axz))
    fig_cube <- fig %>% layout(scene = list(xaxis=axx,yaxis=axy,zaxis=axz, aspectmode='cube')) # To maintain cubic aspect
    return(fig)
}
```


# Reference Data Set(s)

1.  <https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE136719>

2.  <https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE138121>

3.  Rayon 2021 <https://github.com/briscoelab/human_single_cell> 3-1
    <https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE171892> 3-2.
    <https://shiny.crick.ac.uk/scviewer/neuraltube/>

4.  <https://juliendelile.github.io/Antler/>

5.  Yadav 2023
    <https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE190442>
    <https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE222322>

```{bash eval=FALSE}
wget https://ftp.ncbi.nlm.nih.gov/geo/series/GSE190nnn/GSE190442/suppl/GSE190442_aggregated_counts_postqc.csv.gz
wget https://ftp.ncbi.nlm.nih.gov/geo/series/GSE190nnn/GSE190442/suppl/GSE190442_aggregated_metadata_postqc.csv.gz

rclone copy --http-url https://zenodo.org :http:record/4546932 . -P
rclone copy --http-url http://celltypes.brain-map.org :http:api/v2/well_known_file_download/694416044 . -P
rclone copy --http-url http://celltypes.brain-map.org :http:api/v2/well_known_file_download/502175284 . -P
rclone copy refseq:/geo/series/GSE190nnn/GSE190442/suppl/GSE190442_RAW.tar . -P
```

```{bash eval=FALSE}
cd resources/cell_refs/allen_institute/M1

# https://azimuth.hubmapconsortium.org/references/#Human%20-%20Motor%20Cortex
#wget https://seurat.nygenome.org/azimuth/demo_datasets/allen_m1c_2019_ssv4.rds

rclone copy --http-url https://brainmapportal-live-4cc80a57cd6e400d854-f7fdcae.divio-media.net :http:filer_public/70/32/70326830-e306-4743-a02c-a8da5bf9eb56/readme-m1-10.txt  . -P

rclone copy --http-url https://brainmapportal-live-4cc80a57cd6e400d854-f7fdcae.divio-media.net \
  :http:filer_public/0c/0c/0c0c882d-1c31-40a9-8039-3bf2706a77cd/sample-exp_component_mapping_human_10x_apr2020.zip . -P 

rclone copy --http-url https://idk-etl-prod-download-bucket.s3.amazonaws.com :http:aibs_human_m1_10x/metadata.csv . -P
rclone copy --http-url https://idk-etl-prod-download-bucket.s3.amazonaws.com :http:aibs_human_m1_10x/matrix.csv . -P
```

```{bash eval=FALSE}
# https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE171892
cd resources/cell_refs/Rayon_2021
rclone copy refseq:/geo/series/GSE171nnn/GSE171892/suppl . -P

for file in $(ls -1 *.gz)
do
  tar -xzvf $file 
done

cd resources
git clone https://github.com/briscoelab/human_single_cell
```

# Background 

The pMN cell domain gives rise to:
(1) 100s of genetically distinct groups of cholinergic alpha motor neurons clustered into motor pools that innervate specific skeletal muscles; 
(2) gamma motor neurons that innervate intrafusal fibers of specific skeletal muscles for proprioception; 
(3) the predominantly thoracic (T1–12, and to L1 and L2 in some species) cholinergic preganglionic sympathetic neurons;
(4) the cholinergic parasympathetic motor neurons in the sacral (S2–4) cord; and
(5) oligodendrocytes found throughout the spinal cord

# Genome Refs

```{r eval=FALSE}
edb <- ensembldb::EnsDb("resources/genome/Homo_sapiens.GRCh38.98.sqlite")

# edb

gids <- ensembldb::keys(edb, keytype = "GENEID")

idmap <- AnnotationDbi::select(edb, keytype="GENEID", 
                      keys = gids, 
                      columns = c("GENEID","GENEBIOTYPE","GENENAME","SYMBOL")) %>% 
  janitor::clean_names()

head(idmap)
```

```{r eval=FALSE}
allen_genes <- read.csv("resources/cell_refs/allen_institute/human_MTG_2018-06-14_genes-rows.csv")

head(allen_genes)
```

```{r}
marker_genes_mc <- read.csv("resources/cell_refs/allen_institute/allen_institute_azimuth_marker_gene_list.csv") %>% 
  mutate(Label = gsub("\\s{1,}|\\/|-","_", Label)) %>% 
  pull(Markers , name = Label) %>% 
  as.list() %>% 
  sapply(.,  function(x) str_split(gsub("\\.","-",x), pattern = ", "))
```

```{r}
changho_refs_file <- "references/changho_updated_marker_genes_MN_IN_annotation.xlsx"
marker_genes_chun <- purrr::map_dfr(openxlsx::getSheetNames(changho_refs_file)[2:4], function(x){
  openxlsx::read.xlsx(changho_refs_file, sheet = x)
}) %>% 
  janitor::clean_names() %>% 
  distinct() %>% 
  mutate(group=paste(author, tolower(neural_pop)) %>% 
           gsub("\\s","_",.)) %>% 
  mutate(specific_markers_for_each_type_of_neuron=gsub("\\s","", specific_markers_for_each_type_of_neuron))

# marker_genes_chun
marker_genes_chun_list <- purrr::map(unique(marker_genes_chun$group), function(x){
  marker_genes_chun %>%
    filter(group == x) %>%
    pull(specific_markers_for_each_type_of_neuron)
}) %>% 
  set_names(unique(marker_genes_chun$group))
marker_genes_chun_list[["pan_neuronal_markers"]] <- str_split(marker_genes_chun[1,"pan_neuronal_markers"],pattern = ", ", n=3)[[1]]

# marker_genes_chun_list
```

```{r}
# https://github.com/regan-hamel/h-iPSCs-MNI/blob/master/R%20workflow_iMNs_LThiry_2020.Rmd
marker_genes_iPSC <- list(MNPCs = c("CHAT", "ISL1", "ISL2", "MNX1", "OLIG2", "NEUROG2"),
                INs = c("PAX3", "VSX2", "GATA3", "SOX14", "SIM1", "LBX1", "TLX3"),
                NPCs = c("SOX1", "SOX2", "MKI67"),
                Glial = c("SOX9", "S100B", "GFAP"),
                Oligodendrocytes = c("PDGFRA", "GALC"),
                iPSCs = c("NANOG", "POU5F1"))
```

```{r}
marker_genes_endoderm <- c("CXCR4", "KRT12", "SOX17", "KLF8", "MYCT1", "DKK4", "GATA4", "GATA6","SOX7", "EOMES","KRT19","CLDN6","FOXA1","FOXA2")

length(marker_genes_endoderm)
```

<https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4468382/> Cellular
identities in vertebrate spinal cord are specified during development
along the three basic spatial axes of the embryonic body plan \*
rostral--caudal, dorsal--ventral, and medial--lateral.

```{r}
rayon_marker_genes <- openxlsx::read.xlsx(xlsxFile = "resources/cell_refs/Rayon_2021/TableS2.xlsx") %>% 
  janitor::clean_names() %>% 
  mutate(id = paste("rayon", type, neural_pop, sep="_") %>% 
           gsub("\\s", "_", .)) %>% 
  mutate(description = case_when(
    grepl("FP", id) ~ gsub("FP","floor plate", id),
    grepl("RP", id) ~ gsub("RP","roof plate", id),
    grepl("dp", id) ~ gsub("dp[0-9]", "dorsal interneuron progenitors", id),
    grepl("p[0-2]", id) ~ gsub("p[0-2]", "intermediate interneuron progenitors", id),
    grepl("p3", id) ~ gsub("p3","ventral interneuron progenitors p3", id), 
    grepl("V[0-3]", id) ~ gsub("V[0-9].?", "ventral neurons", id),
    grepl("MN", id) ~ gsub(".?MN","motor neuron", id),
    grepl("dl[0-6]", id) ~ gsub("dl[0-9]", "dorsal Interneuron", id),
    grepl("LTMR|LMTR", id) ~ gsub("C?-{0,1}(LTMR|LMTR).{0,}", "mechanoreceptor", id),
    grepl("Proprioceptor|nociceptor", id) ~ gsub("Peripheral","PNS Peripheral",id),
    grepl("Mesoderm|Eryt|Blood|Hema|Myoblast|Oligo", id) ~ paste("rayon",str_split_fixed(id, "_", n=3)[,2]),
    TRUE ~ id
  ) %>% gsub("_"," ", .) %>% gsub("rayon ","", .))


# View(rayon_marker_genes)
head(rayon_marker_genes)
```

```{r}
rayon_marker_genes_long <- rayon_marker_genes %>% 
  separate(genes_map_step1, into = paste0("gene",1:5), sep=", ") %>% 
  separate(genes_map_step2, into = paste0("gene",5:20), sep = ", ") %>% 
  pivot_longer(cols = matches("^gene"),
               names_to = "colname",
               values_to = "genes") %>% 
  filter(!is.na(genes)) 

# rayon_marker_genes_long

rayon_marker_gene_list <-  purrr::map(unique(rayon_marker_genes_long$id), function(x){
  rayon_marker_genes_long %>% 
    filter(id == x) %>% 
    pull(genes) %>% 
    unique()
})
names(rayon_marker_gene_list) <- unique(rayon_marker_genes_long$id)
# rayon_marker_gene_list
```

```{r}
rayon_degs <- openxlsx::read.xlsx("resources/cell_refs/Yadav_2023/1-s2.0-S0896627323000314-mmc5.xlsx", sheet = 1, check.names = FALSE, sep.names = " ") %>% 
  pivot_longer(cols = everything(),
               names_to = "subtype_annotation", 
               values_to = "degs") %>% 
  mutate(names = gsub("-[0-9]{1,2}$","", subtype_annotation) %>% 
           gsub(".+\\s(.+)$", "\\1", .) %>% 
           gsub("\\/|-","_", .))

esc_celltypes <- unique(rayon_degs$names)
rayon_deg_gene_list <-  purrr::map(esc_celltypes, function(x){
  rayon_degs %>%
    filter(names == x) %>%
    pull(degs) %>%
    unique()
})
names(rayon_deg_gene_list) <- esc_celltypes
```

```{r}
allen_ontology <- get_ontology("resources/cell_refs/allen_institute/pcl.obo",
                               extract_tags='everything')
# allen_ontology

human_entries <- data.frame(name = allen_ontology$name, 
                   id = allen_ontology$id,
                   def = allen_ontology$def) %>% 
  filter(!grepl("CHEBI|BFO", id), 
         !grepl("obsolete", name)) %>% 
  filter(grepl("PCL", id), grepl("Hsap", name))

```

```{r}
blum_gamma_mn_degs <- openxlsx::read.xlsx("references/blum_mouse_spinal_cord/41593_2020_795_MOESM3_ESM.xlsx",sheet = "e", startRow = 2) %>% 
  arrange(desc(avg_logFC)) %>% 
  mutate(p_val_adj = as.numeric(p_val_adj)) %>% 
  filter(p_val_adj < 0.001)


blum_alpha_mn_degs <- openxlsx::read.xlsx("references/blum_mouse_spinal_cord/41593_2020_795_MOESM3_ESM.xlsx",sheet = "f", startRow = 2) %>% 
  arrange(desc(avg_logFC)) %>% 
  filter(p_val_adj < 0.001)

blum_gamma_mn_degs #213
blum_alpha_mn_degs #229
```

```{r}
colors_vector <- c(RColorBrewer::brewer.pal(12, "Paired"), 
            "blue1", "darkslategray3", "burlywood3", "#984EA3",
            "seagreen1", "yellow2", "orchid", "darkblue", 
            "lightsalmon2","slateblue1","lightskyblue4",
            "azure2", "chartreuse1",  "lemonchiffon2",
            "deeppink", "darkslategray1", "green4", "navajowhite", 
            "brown4", "darkgoldenrod2", "deepskyblue1", "lightpink") %>% 
  c(., ggpubr::get_palette("jco", 5)) %>% 
  c("mediumpurple4","magenta", "peru") %>% 
  c(ggsci::pal_igv()(51))

length(colors_vector) 
```

# Preprocessing

## Allen Institute:

-   This data set includes single-nucleus transcriptomes from 76,533
    total nuclei derived from 2 post-mortem human brain specimens, to
    survey cell type diversity in the primary motor cortex (M1C or M1).

-   default 10x Cell Ranger v3

-   <https://azimuth.hubmapconsortium.org/references/#Human%20-%20Motor%20Cortex>

-   <https://github.com/satijalab/azimuth-references/tree/master/human_motorcortex>

-   <https://portal.brain-map.org/atlases-and-data/rnaseq/human-m1-10x>

-   <https://knowledge.brain-map.org/data/8GU8FPSNLBUBCJWQDZ5/summary>

-   <https://www.nature.com/articles/s41586-021-03465-8>

Ontologies \*
<https://bioportal.bioontology.org/ontologies/PCL/?p=summary> \*
<https://data.bioontology.org/ontologies/PCL/submissions/21/download?apikey=8b5b7825-538d-40e0-9e9e-5ab9274a9aeb>
\* <https://academic.oup.com/bioinformatics/article/33/7/1104/2843897>

Celltype Explanations:
* https://knowledge.brain-map.org/celltypes/CCN201912131

``` bash
# http://robot.obolibrary.org/convert
robot convert --input pcl.owl --output pcl.obo --check false
```

Code \* <https://github.com/AllenInstitute/BICCN_M1_Evo>

Ages for Allen M1 were reported here:
https://www.biorxiv.org/content/10.1101/2022.10.12.511898v1.supplementary-material
Donor	Age	Sex	Dissection	Dissection abbreviation
H18.30.001	60y	F	Cerebral cortex (Cx) - Precentral gyrus (PrCG) - Primary motor cortex - M1C	M1C
H18.30.002	50y	M	Cerebral cortex (Cx) - Precentral gyrus (PrCG) - Primary motor cortex - M1C	M1C

```{r}
mc_meta <- read.csv("resources/cell_refs/allen_institute/M1/metadata.csv")
rownames(mc_meta) <- mc_meta$sample_name

head(mc_meta)

sample_mapping <- read.csv("resources/cell_refs/allen_institute/M1/sample-exp_component mapping_human_10x_apr2020.csv")

# head(sample_mapping)

az_mc_meta <- read.csv("resources/cell_refs/allen_institute/allen_institute_azimuth_subclass_marker_gene_list.csv") %>% 
  janitor::clean_names()

# head(az_mc_meta)
```

```{r}
celltype_labels <- unique(mc_meta$cluster_label)
length(celltype_labels) #127

allen_defs <- purrr::map_dfr(celltype_labels, function(label){
  human_entries %>% 
    filter(grepl(glue("^{label} "), name), 
           grepl("primary motor cortex", name)) %>% 
    mutate(cluster_label = label)
}) %>% 
  select(cluster_label, everything())

head(allen_defs)
dim(allen_defs)
# celltype_labels[!celltype_labels %in% defs$cluster_label] # missing 2 entries
```

```{r fig.width=10}
cell_tax <- readRDS("resources/cell_refs/allen_institute/M1/human_dendrogram.rds")
plot(cell_tax)
```

```{r}
mc_meta %>% 
  group_by(cluster_label) %>% 
  dplyr::count() %>% 
  arrange(desc(n))

mc_meta %>% 
  group_by(subclass_label) %>% 
  dplyr::count() %>% 
  arrange(desc(n)) %>% 
  write.csv(., "allen_celltypes.csv", row.names = FALSE)
```

```{r eval=FALSE}
ref_mc_counts <- data.table::fread("resources/cell_refs/allen_institute/M1/matrix.csv") 


ref_mc_counts <- as.data.frame(ref_mc_counts)
rownames(ref_mc_counts) <- ref_mc_counts$sample_name
ref_mc_counts <- t(as.matrix(ref_mc_counts[,-1]))

dim(ref_mc_counts) #76533 50282
head(ref_mc_counts[,1:5])
```

```{r eval=FALSE}
ref_motor_cortex <- CreateSeuratObject(counts = ref_mc_counts,
                                   meta.data = mc_meta,
                                   assay="RNA",
                                   project="Allen_M1") 
rm(ref_mc_counts)
rm(mc_meta)
gc()

ref_motor_cortex <- ref_motor_cortex %>% 
  PercentageFeatureSet(
                         assay="RNA",
                         pattern = "^MT-",
                         col.name = "percent.mt") %>% 
  PercentageFeatureSet(
                       assay="RNA",
                       pattern = "^RP[SL]", 
                       col.name = "percent.ribo") %>% 
  SCTransform(
              vst.flavor = "v2", 
              method = "glmGamPoi",
              verbose = T) %>%
    RunPCA(npcs = 50, verbose = T) %>%
    RunUMAP(reduction = "pca", 
            dims = 1:50,
            n.components = 3, 
            min.dist = 0.3,
            verbose = T) %>%
    FindNeighbors(reduction = "pca",
                  dims = 1:50, verbose = T) %>%
    FindClusters(resolution = 0.8, verbose = T)

# saveRDS(ref_motor_cortex,"resources/cell_refs/allen_institute/M1/human_M1_normalized_seurat_object.RDS")
```

```{r}
ref_motor_cortex <- readRDS("resources/cell_refs/allen_institute/M1/human_M1_normalized_seurat_object.RDS")

new_meta <- ref_motor_cortex@meta.data %>% 
  left_join(., az_mc_meta, by=c("cluster_label"="label")) %>% 
  select(sample_name,
         cell_type_designation_label,
         expanded_label, 
         obo_ontology_id,
         cell_type_alias_label, 
         class_label, 
         cluster_label, 
         markers, 
         everything()) %>% 
  set_rownames(.$sample_name)

ref_motor_cortex@meta.data <- new_meta
ref_motor_cortex <- AddMetaData(ref_motor_cortex, metadata = ref_motor_cortex@reductions$umap@cell.embeddings)

# ref_motor_cortex
dim(ref_motor_cortex) #28271 76,533
Idents(ref_motor_cortex) <- ref_motor_cortex$cluster_label
```

```{r eval=FALSE}
ref_motor_cortex@meta.data %>% 
  group_by(obo_ontology_id, expanded_label) %>% 
  dplyr::count() %>% 
  View()

ref_motor_cortex@meta.data %>% 
  group_by(obo_ontology_id, cluster_label) %>% 
  dplyr::count() %>% 
  View()

ref_motor_cortex@meta.data %>% 
  group_by(obo_ontology_id, cell_type_alias_label, class_label) %>% 
  dplyr::count() %>% 
  View()


ref_motor_cortex@meta.data %>% 
  group_by(obo_ontology_id, subclass_label, class_label) %>% 
  dplyr::count() %>% 
  View()
```

```{r}
cluster_color <- ref_motor_cortex@meta.data %>% 
  select(cluster_color, cluster_label) %>% 
  distinct() %>% 
  pull(cluster_color, name = cluster_label)

length(cluster_color)
head(cluster_color)

subclass_color <- ref_motor_cortex@meta.data  %>% 
  select(subclass_label, subclass_color) %>% 
  distinct() %>% 
  pull(subclass_color, name = subclass_label)

table(ref_motor_cortex$external_donor_name_label)
```

```{r fig.height=20, fig.width=20}
# pdf("figures/cell_refs/allen_M1_ref_motor_cortex_cluster_label_UMAP.pdf", height = 20, width = 15)
DimPlot(ref_motor_cortex, 
        repel = TRUE,
        label = TRUE) +
  theme(legend.position = 'bottom') +
  scale_color_manual(values = cluster_color)
# dev.off()
```

```{r fig.height=20, fig.width=20}
# pdf("figures/cell_refs/allen_M1_ref_motor_cortex_subclass_label_UMAP.pdf", height = 20, width = 15)
DimPlot(ref_motor_cortex, 
        group.by = 'subclass_label',
        label.size = 10,
        label.color = 'black',
        label.box = FALSE,
        repel = TRUE,
        label = TRUE) +
  theme(legend.position = 'bottom') +
  scale_color_manual(values = subclass_color)
# dev.off()
```

```{r fig.height=20, fig.width=20}
# pdf("figures/cell_refs/allen_M1_ref_motor_cortex_external_donor_name_label_UMAP.pdf", height = 20, width = 15)
DimPlot(ref_motor_cortex, 
        group.by = "external_donor_name_label",
        label = TRUE) +
  theme(legend.position = 'bottom')
# dev.off()
```

```{r fig.height=20, fig.width=20}
DimPlot(ref_motor_cortex, 
        group.by = "donor_sex_label",
        label = TRUE) +
  theme(legend.position = 'bottom')
```

## Yadav:

fastq files were run through 10x Genomics Cellranger v3.0.2 output bam
files from cellranger were processed using BAMboozle for anonymization

-   Genome_build: hg38 Supplementary_files_format_and_content: Counts
    tables for all barcodes in standard hdf5 output from Cellranger

-   The 7 human datasets were integrated using SCTransform normalization
    followed by CCA based integration using the Seurat

-   <https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM5723843>

```{r}
sc_pts <- read.csv("resources/cell_refs/Yadav_2023/GSE222322_donor_to_patient_ids.csv")


sc_meta <- read.csv("resources/cell_refs/Yadav_2023/GSE190442_aggregated_metadata_postqc.csv") %>% 
  mutate(barcode=X) %>% 
  left_join(., sc_pts, by=c("sample"="Patient_ID")) %>% 
  column_to_rownames("X") 
  
# 
# head(sc_meta)
dim(sc_meta) # 55289     7

sc_pt_meta <- openxlsx::read.xlsx("resources/cell_refs/Yadav_2023/1-s2.0-S0896627323000314-mmc2.xlsx") %>% 
  janitor::clean_names() %>% 
  filter(use=="Sequencing") %>% 
  mutate(donor = as.character(donor))

# head(sc_pt_meta)
dim(sc_pt_meta)
```

```{r}
# Motoneurons	- 1331 only
yada_defs <- sc_meta %>% 
  group_by(top_level_annotation, subtype_annotation) %>% 
  dplyr::count() %>% 
  arrange(desc(n)) %>% 
  ungroup() %>% 
  mutate(description = case_when(
    grepl("Schwann", subtype_annotation) ~ "schwann cells", 
    grepl("OPC", subtype_annotation) ~ "oligodendrocyte precursor cells",
    grepl("Oligo", subtype_annotation) ~ "oligodendrocytes",
    grepl("Micro", subtype_annotation) ~ "microglia",
    grepl("Endo", subtype_annotation) ~ "Venous/Capillary Endothelial and Arterial Endothelial",
    grepl("IEG", subtype_annotation) ~ "neurons expressing immediate-early response genes",
    grepl("Inh-Dorsal", subtype_annotation) ~ "inhibitory dorsal neuron",
    grepl("Ex-Dorsal", subtype_annotation) ~ "excitatory dorsal neurons",
    grepl("Ex-V", subtype_annotation) ~ "excitatory ventral neurons",
    grepl("Ex-M", subtype_annotation) ~ "excitatory mid neurons",
    grepl("Inh-M",subtype_annotation ) ~ "inhibitory mid neurons",
    grepl("Inh-V", subtype_annotation) ~ "inhibitory ventral neurons",
    grepl("Macrophages|Lymphocytes|Motoneurons", subtype_annotation) ~ subtype_annotation,
    TRUE ~ top_level_annotation
  )) %>% 
  mutate_at(vars(description), ~tolower(.)) %>% 
  select(subtype_annotation, description) %>% 
  distinct()
```

Schwann cells (Schwann-1 and Schwann--2) oligodendrocyte precursor cells
(OPCs) progenitors (Oligo Progen) oligodendrocytes (Oligo-1 through
Oligo-6) microglia (Micro-1 through Micro-4 and Perivascular Micro)
macrophages

Astrocytes (WM Astro, GM Astro-1, and GM Astro-2). Meninges (Men-1
through Men-4).

Vascular cells are shown in teal and include two groups of endothelial
cells (Venous/Capillary Endo and Arterial Endo) pericytes/smooth muscle
cells (Peri/SMC) Ependymal cells

Neurons motoneurons (MN), excitatory dorsal neurons (ExDorsal),
inhibitory dorsal neurons (InhDorsal), excitatory mid neurons (ExM),
excitatory ventral neurons (EV), inhibitory mid neurons (InhM),
inhibitory ventral neurons (InhV).

```{r eval=FALSE}
ref_counts_sc <- data.table::fread("resources/cell_refs/Yadav_2023/GSE190442_aggregated_counts_postqc.csv")
ref_counts_sc <- as.data.frame(ref_counts_sc)
rownames(ref_counts_sc) <- ref_counts_sc$V1
ref_mat_sc <- as.matrix(ref_counts_sc[,-1])
head(ref_mat_sc[,1:5])
dim(ref_mat_sc) #22,238 55,290

ref_cell_meta=read.csv("resources/cell_refs/Yadav_2023/GSE190442_aggregated_metadata_postqc.csv", 
                       row.names = 1)
dim(ref_cell_meta) # 55289     6

ref_spinal_cord <- CreateSeuratObject(counts = ref_mat_sc,
                                   meta.data = ref_cell_meta,
                                   assay="RNA",
                                   project="yadav_2023")

ref_spinal_cord

# grep("^MT-", rownames(ref_spinal_cord)) #MT genes are removed?
# grep("^RP[SL]", rownames(ref_spinal_cord), value=TRUE)
# saveRDS(ref_spinal_cord, "resources/cell_refs/Yadav_2023/GSE190442_aggregated_postqc_seurat_obj.RDS")

ref_spinal_cord <- readRDS("resources/cell_refs/Yadav_2023/GSE190442_aggregated_postqc_seurat_obj.RDS") %>% 
  PercentageFeatureSet(
                       assay="RNA",
                       pattern = "^MT-",
                       col.name = "percent.mt") %>% 
  PercentageFeatureSet(
                       assay="RNA",
                       pattern = "^RP[SL]", 
                       col.name = "percent.ribo") %>% 
  SCTransform( 
              vst.flavor = "v2", 
              method = "glmGamPoi",
              verbose = T) %>%
    RunPCA(npcs = 50, verbose = T) %>%
    RunUMAP(reduction = "pca", 
            dims = 1:50,
            n.components = 3, 
            min.dist = 0.3,
            verbose = T) %>%
    FindNeighbors(reduction = "pca",
                  dims = 1:50, verbose = T) %>%
    FindClusters(resolution = 0.8, verbose = T)


ref_spinal_cord <- AddMetaData(ref_spinal_cord, select(sc_meta, barcode:Visium_ID))
table(ref_spinal_cord@meta.data$top_level_annotation)
# saveRDS(ref_spinal_cord, "resources/cell_refs/Yadav_2023/GSE190442_aggregated_postqc_normalized_seurat_obj.RDS")
```

```{r}
ref_spinal_cord <- readRDS("resources/cell_refs/Yadav_2023/GSE190442_aggregated_postqc_normalized_seurat_obj.RDS")

sc_new_meta <- sc_meta %>% 
  left_join(., sc_pt_meta, by=c("sample"="donor")) %>% 
  select(barcode, all_of(colnames(.)[colnames(.) %in% colnames(sc_pt_meta)])) %>% 
  column_to_rownames("barcode")

ref_spinal_cord <- AddMetaData(ref_spinal_cord, sc_new_meta)
Idents(ref_spinal_cord) <- ref_spinal_cord$subtype_annotation
```

```{r fig.height=10, fig.width=10}
# pdf("figures/cell_refs/yadav_GSE190442_ref_spinal_cord_subtype_annotation_UMAP.pdf", height = 20, width = 15)
DimPlot(ref_spinal_cord,
        repel = TRUE,
        label.size = 4,
        label=TRUE) +
  theme(legend.position = 'bottom') +
  scale_color_manual(values=colors_vector)
# dev.off()
```

```{r fig.height=10, fig.width=10}
# pdf("figures/cell_refs/yadav_GSE190442_ref_spinal_cord_top_level_annotation_UMAP.pdf", height = 20, width = 15)
DimPlot(ref_spinal_cord,
        group.by = "top_level_annotation",
        repel = TRUE,
        label.size = 4,
        label=TRUE) +
  theme(legend.position = 'bottom') +
  scale_color_manual(values=colors_vector)
# dev.off()
```

```{r fig.height=10, fig.width=10}
# pdf("figures/cell_refs/yadav_GSE190442_ref_spinal_cord_sample_UMAP.pdf", height = 20, width = 15)
DimPlot(ref_spinal_cord,
        group.by = "sample",
        repel = TRUE,
        label.size = 4,
        label=FALSE) +
  theme(legend.position = 'bottom') +
  scale_color_manual(values=colors_vector)
# dev.off()
```

```{r fig.height=10, fig.width=10}
# pdf("figures/cell_refs/yadav_GSE190442_ref_spinal_cord_sex_UMAP.pdf", height = 20, width = 15)
DimPlot(ref_spinal_cord,
        group.by = "sex",
        repel = TRUE,
        label.size = 4,
        label=FALSE) +
  theme(legend.position = 'bottom') +
  scale_color_manual(values=colors_vector)
# dev.off()
```

## Rayon:

-   10x Genomics Single Cell 3′ Chip V3.1
-   Cell Ranger (3.1.0) and the GRCh38-3.0.0

```{r}
embryonic_sc_meta <- read.csv("references/human_single_cell/metadata/human_metadata.csv") %>% 
  mutate(
        barcode =  X,
        dataset_id = str_split_fixed(X, "_",n=2)[,2]) %>% 
  janitor::clean_names() %>% 
  mutate_at(vars(type_step1, type_step2), ~gsub("\\s{1,}","_", .)) %>% 
  mutate_at(vars(type_step2), ~ifelse(is.na(.), type_step1, .)) %>% 
  column_to_rownames("x")

head(embryonic_sc_meta)
# table(embryonic_sc_meta$orig_timepoint)
```

```{r}
rayon_defs <- rayon_marker_genes %>% 
  select(type, neural_pop, id, description) %>% 
  distinct() %>% 
  mutate_at(vars(type, neural_pop), ~gsub("\\s{1,}","_", .)) %>% 
  mutate_at(vars(neural_pop), ~ifelse(is.na(.), type, .)) 

# rayon_defs

table(rayon_defs$type %in% unique(embryonic_sc_meta$type_step1))
table(rayon_defs$neural_pop %in% unique(embryonic_sc_meta$type_step2))
rayon_defs$neural_pop[!rayon_defs$neural_pop %in% unique(embryonic_sc_meta$type_step2)] #not in manuscript either?
```

```{r}
dataset_suffix <- embryonic_sc_meta %>% 
  group_by(orig_ident, dataset_id) %>% 
  dplyr::count() %>% 
  mutate(dataset_order = gsub("CS12_brachial","CS12_spinalcord", orig_ident) %>% 
           as.factor())

# head(dataset_suffix)
```

```{r}
files <- dir("resources/cell_refs/Rayon_2021/GSE171892_RAW", pattern = "^human_.+", full.names = TRUE) 
# head(files)
```

```{r eval=FALSE}
embryonic_sc_counts <- purrr::map(files, function(x){
  print(x)

  obj <- CreateSeuratObject(counts = Read10X(x),
                            project = basename(x))
  obj <- RenameCells(obj, new.names = gsub("-1","",colnames(obj)))

  return(obj)
})
names(embryonic_sc_counts) <- basename(files)
embryonic_sc_counts <- embryonic_sc_counts[dataset_suffix$dataset_order]
# embryonic_sc_counts$human_CS12_spinalcord_rep1@meta.data
# embryonic_sc_counts$human_CS19_thoracic_rep2@meta.data

# saveRDS(embryonic_sc_counts,"resources/cell_refs/Rayon_2021/GSE171892_seurat_obj_list.RDS")

if(!exists("embryonic_sc_counts")){
  embryonic_sc_counts <- readRDS("resources/cell_refs/Rayon_2021/GSE171892_seurat_obj_list.RDS")
}

idx <- grep("human_CS12_spinalcord_rep1",names(embryonic_sc_counts), invert = TRUE)
embryonic_sc <- merge(x = embryonic_sc_counts[["human_CS12_spinalcord_rep1"]], 
                      y = embryonic_sc_counts[idx])
embryonic_sc #33538 features across 120620 samples within 1 assay
embryonic_sc$barcode <- rownames(embryonic_sc@meta.data)

# head(embryonic_sc@meta.data)
# table(embryonic_sc_meta$barcode %in% embryonic_sc$barcode) #TRUE

all_sc_meta <- embryonic_sc@meta.data %>% 
  mutate(dataset_id = str_split_fixed(barcode, "_",n=2)[,2]) %>% 
  group_by(orig.ident, dataset_id) %>%
  dplyr::count()

# all_sc_meta
# dataset_suffix
# head(rownames(embryonic_sc))
# saveRDS(embryonic_sc, "resources/cell_refs/Rayon_2021/GSE171892_merged_seurat_obj_merged.RDS")
```

**Issue with GSM5236521_human_CS14_thoracic_rep1.tar.gz** --\>
re-downloading this sample resolved the issue Using block 1 from
human_CS14_thoracic_rep1 to learn model. Error in make_cell_attr(umi,
cell_attr, latent_var, batch_var, latent_var_nonreg, : cell attribute
"log_umi" contains NA, NaN, or infinite value

--\> this sample has almost zero read counts 53,967,784 Apr 12 2021
GSM5236521_human_CS14_thoracic_rep1.tar.gz 53,967,784 File
GSM5236521_human_CS14_thoracic_rep1.tar.gz 04/12/2021 09:33:42
53,967,784 TAR

```{r eval=FALSE}
cell_counts <- purrr::map(Layers(embryonic_sc[["RNA"]]),function(x){
  
  cell_totals <- colSums(embryonic_sc[["RNA"]][[x]])
  cell_totals <- cell_totals[order(cell_totals)]
  
})

no_counts <- purrr::map(cell_counts, function(x) which(x == 0))
names(no_counts) <- Layers(embryonic_sc[["RNA"]])


# length(no_counts$counts.human_CS14_thoracic_rep1)
```

```{r eval=FALSE, fig.height=10, fig.width=10}
embryonic_sc <- subset(embryonic_sc, cells = embryonic_sc_meta$barcode)
embryonic_sc <- AddMetaData(embryonic_sc, metadata = embryonic_sc_meta)
# embryonic_sc #33538 features across 43485 samples within 1 assay

embryonic_sc <- embryonic_sc  %>%
  PercentageFeatureSet(
                       assay="RNA",
                       pattern = "^MT-",
                       col.name = "percent.mt") %>%
  PercentageFeatureSet(
                       assay="RNA",
                       pattern = "^RP[SL]",
                       col.name = "percent.ribo")

VlnPlot(embryonic_sc, 
        pt.size = 0,
        features = c("nCount_RNA","nFeature_RNA","percent.mt", "percent.ribo"),
        ncol = 2)

embryonic_sc <- embryonic_sc %>% 
  SCTransform(
              vst.flavor = "v2",
              method = "glmGamPoi",
              verbose = T) %>% 
    RunPCA(npcs = 50, verbose = T)

embryonic_sc <- embryonic_sc %>% 
    RunUMAP(reduction = "pca",
            dims = 1:50,
            n.components = 3,
            min.dist = 0.3,
            verbose = T) %>%
    FindNeighbors(reduction = "pca",
                  dims = 1:50, verbose = T) %>%
    FindClusters(resolution = 0.8, verbose = T)

embryonic_sc$type_step2 <- ifelse(embryonic_sc$type_step2=="", "unknown", embryonic_sc$type_step2)
# saveRDS(embryonic_sc, "resources/cell_refs/Rayon_2021/GSE171892_merged_postqc_normalized_seurat_obj.RDS")
```

```{r}
embryonic_sc_meta %>% 
  group_by(type_step1) %>% 
  dplyr::count() %>% 
  arrange(desc(n)) 
  # write.csv(.,"resources/cell_refs/Rayon_2021/GSE171892_merged_postqc_normalized_seurat_cell_type_table.csv", row.names = FALSE)
```

```{r}
embryonic_sc <- readRDS("resources/cell_refs/Rayon_2021/GSE171892_merged_postqc_normalized_seurat_obj.RDS")

# update the umap params
embryonic_sc <- embryonic_sc %>% 
    RunUMAP(reduction = "pca",
            dims = 1:30,
            n.components = 3,
            min.dist = 0.1,
            verbose = T) %>%
    FindNeighbors(reduction = "pca",
                  dims = 1:50, verbose = T) %>%
    FindClusters(resolution = 0.8, verbose = T)

type_simplified <- embryonic_sc@meta.data %>% 
  mutate(type_simple = case_when(
  grepl("Erythr|Bloo|Hema", type_step1) ~ "Erythrocytes",
  grepl("Mesoderm", embryonic_sc$type_step1) ~ "Mesoderm",
  TRUE ~ type_step1
)) %>% 
  pull(type_simple)

type_step2_simple <- embryonic_sc@meta.data  %>% 
  mutate(type_step2_clean = case_when(
    grepl("^$|Ery|Hema|Meso|Blood",type_step2) ~ type_simple, 
    grepl("Peri",type_simple) ~ type_simple,
    TRUE ~ type_step2)) %>% 
  mutate(type_step2_clean = factor(type_step2_clean, levels = unique(sort(type_step2_clean)))) %>% 
  pull(type_step2_clean) 


# Add Metadata 
embryonic_sc$type_simple <- type_simplified
embryonic_sc$type_step2_simple <- type_step2_simple
```

```{r fig.height=10, fig.width=10}
# pdf("figures/cell_refs/rayon_GSE171892_embryonic_spinal_cord_type_simple_UMAP.pdf", height = 20, width = 15)
DimPlot(embryonic_sc,
        group.by = "type_simple",
        repel = TRUE,
        label.size = 4,
        label=TRUE) +
  theme(legend.position = 'bottom') +
  scale_color_manual(values=colors_vector)
# dev.off()
```


```{r fig.height=10, fig.width=10}
# pdf("figures/cell_refs/rayon_GSE171892_embryonic_spinal_cord_type_step2_UMAP.pdf", height = 20, width = 15)
Idents(embryonic_sc) <- embryonic_sc$type_step2_simple
DimPlot(embryonic_sc,
        repel = TRUE,
        label.size = 4,
        label=TRUE) +
  theme(legend.position = 'bottom') +
  scale_color_manual(values=colors_vector)
# dev.off()
```

```{r fig.height=10, fig.width=10}
# pdf("figures/cell_refs/rayon_GSE171892_embryonic_spinal_cord_orig.ident_UMAP.pdf", height = 20, width = 15)
DimPlot(embryonic_sc,
        group.by = "orig.ident",
        repel = TRUE,
        label.size = 4,
        label=FALSE) +
  theme(legend.position = 'bottom') +
  scale_color_manual(values=colors_vector)
# dev.off()
```

```{r fig.height=10, fig.width=10}
# pdf("figures/cell_refs/rayon_GSE171892_embryonic_spinal_cord_orig_tissue_UMAP.pdf", height = 20, width = 15)
DimPlot(embryonic_sc,
        group.by = "orig_tissue",
        repel = TRUE,
        label.size = 4,
        label=FALSE) +
  theme(legend.position = 'bottom') +
  scale_color_manual(values=colors_vector)
# dev.off()
```


# Integrate

un-normalized counts (layer='counts'), normalized data (layer='data'),
z-scored/variance-stabilized data (layer='scale.data')

<https://github.com/satijalab/seurat/issues/7587> if using merge() \>
Error in DelayedArray::colAutoGrid(x = counts, ncol = length(x =
layer.cells)) : \> 'ncol' is too big. Blocks of length \>
.Machine\$integer.max are not supported yet. Please specify a smaller
'ncol'. merged_obj \<- merge(x = ref_spinal_cord, y = ref_motor_cortex,
merge.data = FALSE)

### Merge Objects

```{r eval=FALSE}
if(!exists("ref_motor_cortex")){
  ref_motor_cortex  <- readRDS("resources/cell_refs/allen_institute/M1/human_M1_normalized_seurat_object.RDS")
  DefaultAssay(ref_motor_cortex) <- "RNA"
}

if(!exists("ref_spinal_cord")){
  ref_spinal_cord <- readRDS("resources/cell_refs/Yadav_2023/GSE190442_aggregated_postqc_normalized_seurat_obj.RDS")
  DefaultAssay(ref_spinal_cord) <- "RNA"
}

if(!exists("embryonic_sc")){
  embryonic_sc <- readRDS("resources/cell_refs/Rayon_2021/GSE171892_merged_postqc_normalized_seurat_obj.RDS")
  DefaultAssay(embryonic_sc) <- "RNA"
}
```

```{r eval=FALSE}
# https://www.biostars.org/p/388416/
rnames <- sapply(list(ref_motor_cortex, ref_spinal_cord, embryonic_sc), rownames)
lapply(rnames, head, n=30)
sapply(rnames, length)
purrr::reduce(rnames, intersect) %>% length() # 20,243 gene names in common in RNA assay
```

```{r eval=FALSE}
# There is a LARGE difference in mean counts per cell in Allen Institute dataset. 
# https://github.com/satijalab/seurat/issues/5340
# https://github.com/satijalab/seurat/issues/5009
cell_read_depth <- list(cell_read_depth_mc = colSums(ref_motor_cortex[["RNA"]]$counts), 
                        cell_read_depth_sc = colSums(ref_spinal_cord[["RNA"]]$counts), 
                        cell_read_depth_esc = colSums(JoinLayers(embryonic_sc[["RNA"]])$counts))

# saveRDS(cell_read_depth,"results/integrated_ref/read_depth_per_cell_and_dataset.RDS")
sapply(cell_read_depth, mean)
```

```{r eval=FALSE}
ref_spinal_cord$donor <- paste0("sample.", ref_spinal_cord$sample)
ref_motor_cortex$donor <- ref_motor_cortex$external_donor_name_label
embryonic_sc$donor <- embryonic_sc$orig_timepoint

# top_level_annotation == subclass_label, type_step1
# subtype_annotation == cluster_label, type_step2
merged_meta <- ref_spinal_cord@meta.data %>% 
  bind_rows(., 
            select(ref_motor_cortex@meta.data, 
                   barcode = sample_name,
                   subtype_annotation = cluster_label ,
                   top_level_annotation = subclass_label, 
                   everything())) %>% 
  bind_rows(., select(embryonic_sc@meta.data,
                      subtype_annotation = type_step2,
                      top_level_annotation = type_step1,
                      everything())) %>% 
  select(barcode, orig.ident:subtype_annotation, donor, everything())

# head(merged_meta)
dim(merged_meta) # 175,307
table(merged_meta$donor)
# write.csv(merged_meta, "results/integrated_ref/Allen_Rayon_Yadav_integrated_cell_ref_metadata.csv", row.names = F)

merged_obj <-  CreateSeuratObject(counts = list("ref_spinal_cord"=LayerData(ref_spinal_cord, layer = 'counts'),
                                                "ref_motor_cortex"=LayerData(ref_motor_cortex, layer = 'counts'),
                                                "ref_embryonic_sc"=LayerData(JoinLayers(embryonic_sc), layer = 'counts')), 
                                  meta.data = merged_meta)

merged_obj <- JoinLayers(merged_obj)

merged_obj
rm(ref_spinal_cord, ref_motor_cortex, embryonic_sc)
# table(merged_obj$donor)
# saveRDS(merged_obj, "resources/cell_refs/integrated/embryonic_adult_spinal_cord_motor_cortex_merged_seurat_obj.RDS")
```

## Merge and Clean Cell Annotations

```{r eval=FALSE}
if(!exists("merged_obj")){
  merged_obj <- readRDS("resources/cell_refs/integrated/embryonic_adult_spinal_cord_motor_cortex_merged_seurat_obj.RDS")
}

# Split by sample donor
DefaultAssay(merged_obj) <- "RNA"
merged_split_obj <-  split(merged_obj, f = merged_obj$donor) 
rm(merged_obj)
gc()

# remove MT genes for consistency across datasets as some retained these, others removed them
merged_split_obj <- merged_split_obj[-grep("^MT-", rownames(merged_split_obj)),]

# SCT normalization
merged_split_obj <- merged_split_obj %>% 
  SCTransform(method = "glmGamPoi",
              vst.flavor = "v2", 
              verbose = TRUE) %>% 
  RunPCA(npcs = 50,
         verbose = TRUE)

# saveRDS(merged_split_obj,"results/integrated_ref/embryonic_adult_spinal_cord_motor_cortex_merged_split_normalized_seurat_obj.RDS")
```

```{r eval=FALSE}
if(!exists("merged_split_obj")){
  merged_split_obj <- readRDS("results/integrated_ref/embryonic_adult_spinal_cord_motor_cortex_merged_split_normalized_seurat_obj.RDS")
}

if(!exists("sc_pt_meta")){
  sc_pt_meta <- openxlsx::read.xlsx("resources/cell_refs/Yadav_2023/1-s2.0-S0896627323000314-mmc2.xlsx") %>% 
    janitor::clean_names() %>% 
    filter(use=="Sequencing") %>% 
    mutate(donor = as.character(donor))
}

# Add Meta Data
merged_split_obj$dataset <- gsub(".+_CS[0-9]{2}_.+","Rayon_2021",merged_split_obj$orig.ident)
merged_split_obj$top_level_annotation <- gsub("\\s{1,}","_",merged_split_obj$top_level_annotation)
merged_split_obj$subtype_annotation <- gsub("\\s{1,}","_",merged_split_obj$subtype_annotation)
merged_split_obj$motorneuron <- ifelse(grepl("Moto|MN|pMN",merged_split_obj$subtype_annotation), merged_split_obj$subtype_annotation, "Other")

#CS12 and CS17 indicating these samples were male, others female
sc_meta_to_add <- merged_split_obj@meta.data %>% 
  select(-barcode) %>% 
  rownames_to_column("barcode") %>% 
  mutate(sample = as.character(sample)) %>% 
  left_join(., sc_pt_meta, by=c("sample"="donor")) %>% 

  # https://www.ehd.org/virtual-human-embryo/ages.php?stage=1
  mutate(age_clean = case_when(
    dataset == "yadav_2023" ~ paste(age, "yrs", sep="_"),
    grepl("CS12", orig.ident) ~ "29-31_days",
    grepl("CS14", orig.ident) ~  "33-35_days",
    grepl("CS17", orig.ident) ~ "39-42_days",
    grepl("CS19", orig.ident) ~ "45-47_days",
    TRUE ~ "unknown"), 
    sex_clean = case_when(
      dataset == "yadav_2023" ~ sex, 
      dataset == "Allen_M1" ~ donor_sex_label,
      grepl("CS12|CS17", orig.ident) ~ "M",
      grepl("CS14|CS19", orig.ident) ~ "F",
      TRUE ~ "unknown"
    )) %>% 
  select(barcode, age_clean, sex_clean, tissue, cause_of_death) %>% 
  set_rownames(.$barcode)

# Add metadata to object
merged_split_obj <- AddMetaData(merged_split_obj, metadata = sc_meta_to_add)
merged_split_obj <- clean_standardize_annots(merged_split_obj)

DefaultAssay(merged_split_obj)
merged_split_obj
```

```{r eval=FALSE}
all_defs <- allen_defs %>%
  select(subtype_annotation=cluster_label,
         description = def) %>%
  bind_rows(., yada_defs) %>%
  bind_rows(., select(rayon_defs,
                      subtype_annotation = neural_pop,
                      description)) %>%
  mutate(subtype_annotation = gsub("\\s{1,}","_",subtype_annotation))

top_level_table_merged <- merged_split_obj@meta.data %>%
  group_by(dataset,class_label, top_level_annotation, subtype_annotation) %>%
  dplyr::count(name = "number_of_cells") %>%
  arrange(top_level_annotation) %>%
  left_join(.,all_defs, by="subtype_annotation") %>%
  mutate_at(vars(description), ~gsub('\\"', "", .))

# top_level_table_merged
# write.csv(top_level_table_merged, "results/integrated_ref/Allen_Rayon_Yadav_integrated_celltype_labels_table.csv", row.names = FALSE)
```

We next assigned putative neurotransmitter status to each population,
identifying 19 glutamatergic populations (defined by the expression of
SLC17A6) and 14 GABA/glycinergic populations (defined by expression of
GAD1, GAD2, PAX2, and SLC6A5).

```{r}
merged_split_obj@meta.data %>% 
  group_by(dataset, top_level_standardized) %>% 
  dplyr::count()

merged_split_obj@meta.data %>% 
  group_by(subtype_standardized) %>% 
  dplyr::count() %>% 
  View()
```


### Integrate Layers


SCT integration:

- <https://github.com/satijalab/seurat/issues/7542>

- <https://github.com/satijalab/seurat/issues/7145> 
- You don't need to worry about those warnings. We will fix them in Seurat-object package.
- Warning: Different cells and/or features from existing assay SCT

--\> Also, did not use plan(multisession) though unsure if that was
actually causing the errors.


Note on theta from Harmony:

https://www.nature.com/articles/s41592-019-0619-0
As θ increases, the objective function favors more independence between batch f and cluster assignment. As θ approaches infinity, it will yield a degenerate solution. In this case, each cluster has an equivalent distribution across batch f.

theta	
Diversity clustering penalty parameter. Specify for each variable in group.by.vars. Default theta=2. theta=0 does not encourage any diversity. Larger values of theta result in more diverse clusters.

lambda	
Ridge regression penalty parameter. Specify for each variable in group.by.vars. Default lambda=1. Lambda must be strictly positive. Smaller values result in more aggressive correction.

```{r}
# clean and standardized celltypes to include in the reference
celltype_annots <- read.csv("references/Allen_Rayon_Yadav_integrated_celltype_labels_table_CCedit.csv") %>% 
  select(dataset, matches("standard"), everything()) %>% 
  mutate(standardized_top_level_annotation = gsub("\\s{1,}", "_", standardized_top_level_annotation)) %>% 
  mutate_at(vars(standardized_top_level_annotation), ~case_when(
    dataset == "yadav_2023" & top_level_annotation == "Neurons" ~ top_level_annotation,
    TRUE ~ .))

celltypes_to_retain <- celltype_annots %>% 
  filter(standardized_top_level_annotation != "Remove", 
         !grepl("Ery", top_level_annotation), 
         !grepl("Null", subtype_annotation)) %>% 
  mutate_at(vars(standardized_subtype_annotation), ~gsub("\\s{1,}","_", .) %>%
              # gsub("(Neuron)s","\\1",.) %>% 
              ifelse(grepl("Not|Rem", .), standardized_top_level_annotation, .))

# head(celltypes_to_retain)
dim(celltypes_to_retain)

sum(celltypes_to_retain$number_of_cells) # 87,603
table(celltypes_to_retain$standardized_top_level_annotation)
```

```{r eval=FALSE}
params <- list("v1"=list(k_anchor = 5, theta = NULL, lambda = NULL, subset = FALSE),
               "v2"=list(k_anchor = 50, theta = 4, lambda = NULL, subset = FALSE),
               "v3"=list(k_anchor = 20, theta = 0.5, lambda = NULL, subset = FALSE),
               "v4"=list(k_anchor = 20, theta = NULL, lambda = NULL, subset = TRUE))

# Integration
integration_jobs <-  purrr::map(names(params)[c(4)], function(param_set) {
  input_params <- params[[param_set]]
  print(c("Using params: ", input_params))

  if(input_params[["subset"]]){
    print("Subsetting the seurat object")
    idx <- which(merged_split_obj$subtype_annotation %in% celltypes_to_retain$subtype_annotation)
    subset_meta_data <- merged_split_obj@meta.data[idx,]
    DefaultAssay(merged_split_obj) <- "RNA"
    merged_split_obj <- subset(merged_split_obj, cells = rownames(subset_meta_data)) %>% 
        # re-normalize the RNA assay on the subset of cells
        SCTransform(method = "glmGamPoi",
              vst.flavor = "v2", 
              verbose = TRUE) %>% 
        RunPCA(npcs = 50, verbose = TRUE)
    print(dim(merged_split_obj))
  }

  # Background the processes
  job::job({
    Idents(merged_split_obj) <- merged_split_obj$donor
    # k.anchor = 5 is default used here
    merged_split_obj <- IntegrateLayers(merged_split_obj,
                                  method = RPCAIntegration,
                                  orig.reduction = "pca",
                                  new.reduction = "integrated.rpca",
                                  normalization.method = 'SCT', #why is this required?
                                  assay = "SCT",
                                  k.anchor = input_params[["k_anchor"]],
                                  verbose = TRUE)
    # harmony groups are the 13 SCT models (1 per donor)
    # default theta = NULL
    merged_split_obj <- IntegrateLayers(merged_split_obj,
                                  method = HarmonyIntegration,
                                  orig.reduction = "pca",
                                  new.reduction = "harmony",
                                  assay = "SCT",
                                  theta = input_params[["theta"]],
                                  lambda = input_params[["lambda"]],
                                  verbose = TRUE)

    # Run UMAP
    merged_split_obj <- merged_split_obj %>%
      RunUMAP(n.components = 3L,
                   reduction = "pca",
                   dims = 1:50,
                   reduction.name = "umap.unintegrated") %>%
      RunUMAP(n.components = 3L,
                   reduction = "harmony",
                   dims = 1:50,
                   reduction.name = "umap.harmony") %>%
      RunUMAP(n.components = 3L,
                  reduction = "integrated.rpca",
                   dims = 1:50,
                   reduction.name = "umap.rpca")

    # save object
    saveRDS(merged_split_obj,
            glue("results/integrated_ref/embryonic_adult_spinal_cord_motor_cortex_integrated_normalized_seurat_obj_{param_set}.RDS"))
    rm(merged_split_obj)
    gc()
    input_params
    
  }, import = c("merged_split_obj", "input_params", "param_set"))
})
# rm(merged_split_obj)
```

https://satijalab.org/seurat/articles/seurat5_integration.html
-   You will need to do rejoin the layers before performing any
    differential expression analysis.
-   However, you can always resplit the layers in case you would like to
    reperform integrative analysis.

### Sketch Integration

importantly, the estimation of leverage scores only **requires data normalization**, can be computed efficiently for sparse datasets, and does not require any intensive computation or dimensional reduction steps.

```{r eval=FALSE}
if(!exists("merged_split_obj")){
  merged_split_obj <- readRDS("results/integrated_ref/embryonic_adult_spinal_cord_motor_cortex_merged_split_normalized_seurat_obj.RDS")
}

DefaultAssay(merged_split_obj) <- "RNA"
merged_sketch_obj <- DietSeurat(merged_split_obj, assays = "RNA")
merged_sketch_obj <- NormalizeData(merged_sketch_obj) #log-normalize for this proceedure. Does not appear to support SCT. 
merged_sketch_obj <- FindVariableFeatures(merged_sketch_obj)
merged_sketch_obj <- SketchData(object = merged_sketch_obj,
                               ncells = 10000, 
                               assay = "RNA",
                               method = "LeverageScore", 
                               sketched.assay = "sketch")

merged_sketch_obj
```

Harmony: theta\
Diversity clustering penalty parameter. Specify for each variable in
vars_use Default theta=2. theta=0 does not encourage any diversity.
Larger values of theta result in more diverse clusters.

```{r eval=FALSE}
DefaultAssay(merged_sketch_obj) <- "sketch"

merged_sketch_obj <- merged_sketch_obj %>% 
  FindVariableFeatures() %>% 
  ScaleData() %>% 
  RunPCA(npcs = 50, 
         reduction.name = "sketch.pca",
         verbose = TRUE)

# integrate the datasets
job::job
merged_sketch_obj <- IntegrateLayers(merged_sketch_obj, 
                              method = RPCAIntegration,
                              orig.reduction = "sketch.pca",
                              new.reduction = "sketch.rpca", 
                              k.anchor = 50,
                              n.trees = 100,
                              verbose = TRUE)

merged_sketch_obj <- IntegrateLayers(merged_sketch_obj,
                              method = HarmonyIntegration,
                              orig.reduction = "sketch.pca",
                              theta = 4,
                              new.reduction = "sketch.harmony", #name must be hardcoded in source code - it will NOT change no matter what. 
                              verbose = TRUE)

# Run UMAP
merged_sketch_obj <- RunUMAP(merged_sketch_obj,
               reduction = "sketch.pca",
               dims = 1:50,
               reduction.name = "umap.unintegrated") 

merged_sketch_obj <- merged_sketch_obj %>% 
  RunUMAP( reduction = "harmony",
               dims = 1:50,
               reduction.name = "umap.harmony") %>% 
  RunUMAP( reduction = "sketch.rpca",
               dims = 1:50,
               reduction.name = "umap.rpca")

merged_sketch_obj

job::job({
    saveRDS(merged_sketch_obj,"results/integrated_ref/embryonic_adult_spinal_cord_motor_cortex_integrated_normalized_sketch_seurat_obj.RDS")
}, import = c("merged_sketch_obj"))
```

```{r}
merged_sketch_obj <- readRDS("results/integrated_ref/embryonic_adult_spinal_cord_motor_cortex_integrated_normalized_sketch_seurat_obj.RDS")
merged_sketch_obj
```

```{r}
updated_labels <- read.csv("results/integrated_ref/updated_cell_type_labels.csv", row.names = 1)
merged_sketch_obj <- AddMetaData(merged_sketch_obj, updated_labels)

age_tissue_clean <- merged_sketch_obj@meta.data %>% 
  mutate(tissue_clean = case_when(
  dataset == "Allen_M1" ~ "adult_motor_cortex",
  dataset == "Rayon_2021" ~  paste0(gsub("human_|_rep[0-9]|spinalcord","", orig.ident), "_spinal_cord"),
  dataset == "yadav_2023" ~ gsub("adult_Lumbar_spinal_cord_", "adult", tolower(tissue))), 
  
  age_clean = case_when(
    donor == "H18.30.002" ~ "50_yrs",
    donor == "H18.30.001" ~ "60_yrs",
    TRUE ~ gsub("\\~", "", age_clean))) %>% 
  select(age_clean, tissue, tissue_clean)

merged_sketch_obj$tissue <- age_tissue_clean$tissue_clean
merged_sketch_obj$age_clean <- age_tissue_clean$age_clean

table(merged_sketch_obj$age_clean)
table(merged_sketch_obj$tissue)
```

```{r}
merged_sketch_obj@meta.data %>% 
  group_by(top_level_standardized, dataset) %>% 
  dplyr::count() %>% 
  pivot_wider(names_from = dataset, values_from = n) %>% 
  mutate_at(vars(Allen_M1:Rayon_2021), ~ifelse(is.na(.), 0, .)) %>% 
  write.csv(., "cell_types_per_dataset.csv",row.names = FALSE)
```

```{r}
colors_mn <- c("Other"="grey80", "Motoneurons" = "#641A80FF", pMN = "#B63679FF" ,MN = "#FECE91FF")

reds <- c("umap.unintegrated", "umap.harmony", "umap.rpca")
meta_cols <- c("motorneuron", "age_clean", "donor", "tissue","top_level_annotation", "top_level_standardized", "class_label_standardized", "subtype_standardized")
to_plot <- crossing(reds, meta_cols)

purrr::map(1:nrow(to_plot), function(x){
  to_plot <- to_plot %>% 
    dplyr::slice(x)
  red <- to_plot[["reds"]]
  col <- to_plot[["meta_cols"]]
  
  outfile <- glue("figures/cell_refs/allen_yadav_rayon_sketch/embryonic_adult_spinal_cord_motor_cortex_{gsub('umap.','', red)}_sketch_{col}_UMAP.pdf")
  cc <- ifelse(grepl("motorneuron", col), "colors_mn", "colors_vector")
  dp <- DimPlot(merged_sketch_obj,
        group.by = col,
        reduction = red,
        label = TRUE,
        repel = TRUE) +
    scale_color_manual(values = get(cc)) +
    theme(legend.position = "bottom")
  
  pdf(outfile, height = 10, width = 11)
  print(dp)
  dev.off()
})
```

## SCANVI 

When our dataset is fully labelled (perhaps in independent studies, or independent analysis pipelines), we can obtain an integration that better preserves biology using scANVI, which incorporates cell type annotation information.
https://docs.scvi-tools.org/en/stable/tutorials/notebooks/scrna/harmonization.html

all of our data is annotated. 
This can lead to a more accurate integration result when using scANVI
scANVI requires:
  * the sample identifier for each cell (as in scVI)
  * the cell type/state for each cell

```{bash eval=FALSE}
mamba env create -f env/scanvi.yaml
```

```{r message = FALSE, warning=FALSE}
library(sceasy)
library(anndata)
# FutureWarnings
reticulate::use_condaenv(condaenv = file.path(Sys.getenv("HOME"), "opt/mambaforge/envs/scanvi_scib"))

py_config()
# python:         /Users/jsmi26/opt/mambaforge/envs/scanvi/bin/python
# libpython:      /Users/jsmi26/opt/mambaforge/envs/scanvi/lib/libpython3.10.dylib
# pythonhome:     /Users/jsmi26/opt/mambaforge/envs/scanvi:/Users/jsmi26/opt/mambaforge/envs/scanvi
# version:        3.10.13 | packaged by conda-forge | (main, Oct 26 2023, 18:09:17) [Clang 16.0.6 ]
# numpy:          /Users/jsmi26/opt/mambaforge/envs/scanvi/lib/python3.10/site-packages/numpy
# numpy_version:  1.24.4

sc <- import("scanpy", convert = FALSE)
scvi <- import("scvi", convert = FALSE)
```

https://docs.scvi-tools.org/en/stable/tutorials/notebooks/quick_start/api_overview.html

* Unless otherwise specified, scvi-tools models require the raw counts (not log library size normalized)
* perform feature selection, to reduce the number of features (genes in this case) used as input to the scvi-tools model
* we recommend anywhere from 1,000 to 10,000 HVGs, but it will be context-dependent.

```{r}
if(!exists("merged_split_obj")){
  # start with the subsetted dataset
  merged_split_obj <- readRDS("results/integrated_ref/embryonic_adult_spinal_cord_motor_cortex_integrated_normalized_seurat_obj_v4.RDS")
}

# Note: scvi-tools requires raw gene expression
DefaultAssay(merged_split_obj) <- "RNA"
merged_scanvi_obj <- DietSeurat(merged_split_obj, assays = "RNA")
merged_scanvi_obj <- clean_standardize_annots(merged_scanvi_obj)

# re-join layers since convertFormat expects a Seurat v3 object, not v5
merged_scanvi_obj <- JoinLayers(merged_scanvi_obj)
merged_scanvi_obj[["RNA"]] <- as(object = merged_scanvi_obj[["RNA"]], Class = "Assay")
# log-normalize for finding variable genes 
merged_scanvi_obj <- NormalizeData(merged_scanvi_obj)  
merged_scanvi_obj <- FindVariableFeatures(merged_scanvi_obj)


# merged_scanvi_obj
# table(merged_scanvi_obj$subtype_standardized)
# table(merged_scanvi_obj$donor)
# table(merged_scanvi_obj$dataset)

# remove large objs from memory
rm(merged_split_obj)
```

```{r}
cell_type_table <- merged_scanvi_obj@meta.data %>% 
  group_by(dataset, subtype_standardized) %>% 
  dplyr::count() %>% 
  pivot_wider(names_from = dataset, values_from = n) %>% 
  mutate(across(Allen_M1:yadav_2023, ~ifelse(is.na(.), 0,.)))

# write.csv(cell_type_table, "results/integrated_ref/Allen_Rayon_Yadav_integrated_standardized_celltype_labels_table.csv", row.names = FALSE)
```

https://anndata.readthedocs.io/en/latest/generated/anndata.AnnData.html

AnnData format usually contains the following slots:
- X contains the expression matrix.
- obsm contains the embeddings data.
- obs contains the cell metadata.
- var contains the gene metadata.
- varm: Multi-dimensional annotation of variables/features (mutable structured ndarray).
- uns Key-indexed unstructured annotation.


Data structure example:
--> n_obs == cells as rows
--> n_vars == genes as columns
AnnData object with n_obs × n_vars = 32472 × 15148
    obs: 'dataset', 'location', 'nGene', 'nUMI', 'patientGroup', 'percent.mito', 'protocol', 'sanger_type', 'size_factors', 'sampling_method', 'batch', 'cell_type', 'donor'
    layers: 'counts'
    
```{r}
# 34972 genes features across 87,603 cells
# transpose the raw counts matrix
# note: MT genes have already been removed from the counts matrix
tr_counts_mat <- t(as.matrix(GetAssayData(merged_scanvi_obj, layer = "counts")))

# convert to anndata object
adata <- AnnData(
  X = tr_counts_mat, # not standard to put raw counts in X, but I want to ensure only raw counts are used in all steps. 
  obs = as.data.frame(merged_scanvi_obj@meta.data),
  var = as.data.frame(merged_scanvi_obj@assays$RNA@meta.features),
  layers = list(counts = tr_counts_mat)
)

# save another copy of the raw dataset in the object 
adata$raw = adata
adata
class(adata) #"AnnDataR6" "R6"
rm(tr_counts_mat)
rm(merged_scanvi_obj)
```

--> missing layers in the output with convertFormat
adata <- convertFormat(merged_scanvi_obj,
                       from="seurat",
                       to="anndata",
                       main_layer="counts",
                       drop_single_values=FALSE)
print(adata)
--> adata is transposed here as well
AnnData object with n_obs × n_vars = 87603 × 34972

batch_key
  - If specified, highly-variable genes are selected within each batch separately and merged. This simple process avoids the selection of batch-specific genes and acts as a lightweight batch correction method.
  
```{r}
# verbosity to max
sc$settings$verbosity = 4L

# retain genes expressed in at least 10 cells
sc$pp$filter_genes(adata, min_cells = 10)
# filtered out 9727 genes that are detected in less than 10.0 cells
dim(adata) # 87603 by 25245 genes

sc$pp$highly_variable_genes(
    adata,
    n_top_genes=5000,
    subset=TRUE,
    layer="counts",
    flavor="seurat_v3",
    batch_key="dataset",
)

dim(adata) # 87603 by  5000 genes
# adata
```

FutureWarning: SparseDataset is deprecated and will be removed in late 2024. It has been replaced by the public classes CSRDataset and CSCDataset.

```{r}
# set seed for randomization 
scvi$settings$seed <- as.integer(20231030)

# verbosity < 20 is INFO and verbosity < 10 is DEBUG
# scvi$settings$verbosity 
# run setup_anndata, use column  dataset for batch and donor 
# https://scanpy.readthedocs.io/en/stable/generated/scanpy.pp.highly_variable_genes.html
scvi$model$SCVI$setup_anndata(adata,
                              layer="counts", 
                              categorical_covariate_keys = c('dataset','donor'))

# adata
```

Variational Inference:
https://www.nature.com/articles/s41592-018-0229-2
https://docs.scvi-tools.org/en/stable/api/reference/scvi.model.SCVI.html#scvi.model.SCVI

Params for benchmarking study
https://github.com/chenlingantelope/HarmonizationSCANVI/blob/master/notebooks/harmonization-CitePure-SCANVI.ipynb
https://github.com/theislab/scib-reproducibility/blob/main/notebooks/integration/scanvi.ipynb

Dispersion:

'gene' - dispersion parameter of NB is constant per gene across cells

'gene-batch' - dispersion can differ between different batches

'gene-label' - dispersion can differ between different labels

'gene-cell' - dispersion can differ for every gene in every cell

```{r, eval=FALSE}
# min(round((20000 / adata$shape[1]) * 400), 400) # 91 max_epochs by default for training

# create the model
# single-cell Variational Inference params recommended by authors
# n_layers=2, # Number of hidden layers used for encoder and decoder NNs
# n_latent=30, # Dimensionality of the latent space
# gene_likelihood="nb" #Negative binomial distribution
model = scvi$model$SCVI(adata,  
                        n_layers=2L, 
                        n_latent=30L, 
                        dispersion = 'gene', 
                        gene_likelihood="nb")
model
# adata

# Add multiple threads
scvi$settings$dl_num_workers = 2L
scvi$settings$num_threads = 50L
scvi$settings$batch_size = 1024L

# train the model
# https://docs.scvi-tools.org/en/stable/api/reference/scvi.model.SCVI.html#scvi.model.SCVI.train
# 0.9 train_size is default
# increase to 400 epochs for increased accuracy
# it take 6hrs 45min to complete training! 
model$train(accelerator = 'cpu', 
            devices = 'auto',
            max_epochs = 400L,
            train_size = 0.9)

# Saving consists of saving the model neural network weights, as well as parameters used to initialize the model.
model_dir <- "results/integrated_ref/scanvi"
model$save(dir_path = model_dir, prefix = 'embryonic_adult_spinal_cord_motor_cortex', save_anndata = TRUE, overwrite=TRUE)
model
```

```{r warning=FALSE, message=FALSE}
# reading in the adata obj using reticulate changes the data structure/class. 
#sc$read_h5ad(path) #"anndata._core.anndata.AnnData" "python.builtin.object"
adata <- anndata::read_h5ad("results/integrated_ref/scanvi/embryonic_adult_spinal_cord_motor_cortexadata.h5ad")
# class(adata)

SCVI_NORMALIZED_KEY <- "scvi_normalized"
SCVI_LATENT_KEY <- "X_scVI"
norm_cts <- model$get_normalized_expression(library_size=10e4)
latent <- model$get_latent_representation()

# unable to assign obsm in reticulate/anndataR in the same way as adata$layers. Limitation of the package currently. 
# https://github.com/theislab/zellkonverter/issues/75
adata_update <- AnnData(
  X = adata$X,
  obs = adata$obs, 
  var = adata$var, 
  uns = adata$uns,
  obsm = list("_scvi_extra_categorical_covs" = adata$obsm$`_scvi_extra_categorical_covs`,
              "X_scVI" = latent), 
  layers = list("counts" = adata$layers['counts'],
                "scvi_normalized" = norm_cts)
)

adata_update

# run PCA then generate UMAP plots
sc$tl$pca(adata_update)
sc$pp$neighbors(adata_update, n_pcs=30L, n_neighbors=20L)
sc$tl$umap(adata_update, min_dist=0.3)
```

```{r fig.height=10, fig.width=10}
# pdf("figures/cell_refs/allen_yadav_rayon_integrated/subset/scanvi/scanvi_subtype_standardized_unintegrated_UMAP.pdf", height = 15, width = 20)
sc$pl$umap(
    adata_update,
    color=c("subtype_standardized"),
    frameon=TRUE,
)
# dev.off()

# pdf("figures/cell_refs/allen_yadav_rayon_integrated/subset/scanvi/scanvi_dataset_donor_unintegrated_UMAP.pdf", height = 10, width = 20)
sc$pl$umap(
    adata_update,
    color= c('dataset','donor'),
    ncols=2L,
    frameon=TRUE,
)
# dev.off()
```

```{r}
# use scVI latent space for UMAP generation
#  We see that this dataset is quite complex, where only some batches contain certain cell types.
sc$pp$neighbors(adata_update, use_rep=SCVI_LATENT_KEY)
sc$tl$umap(adata_update, min_dist=0.3)

# pdf("figures/cell_refs/allen_yadav_rayon_integrated/subset/scanvi/scanvi_subtype_standardized_scvi_UMAP.pdf", height = 15, width = 20)
sc$pl$umap(
    adata_update,
    color=c("subtype_standardized"),
    frameon=TRUE,
)
# dev.off()

# pdf("figures/cell_refs/allen_yadav_rayon_integrated/subset/scanvi/scanvi_dataset_donor_scvi_UMAP.pdf", height = 10, width = 20)
sc$pl$umap(
    adata_update,
    color=c('dataset','donor'),
    ncols=2L,
    frameon=TRUE,
)
# dev.off()

# pdf("figures/cell_refs/allen_yadav_rayon_integrated/subset/scanvi/scanvi_motorneuron_scvi_UMAP.pdf", height = 10, width = 12)
sc$pl$umap(
    adata_update,
    color=c('motorneuron'),
    frameon=TRUE,
)
# dev.off()
```

n_samples_per_label (Optional[float] (default: None)) – Number of subsamples for each label class to sample per epoch. By default, there is no label subsampling.

```{r eval=FALSE}
scanvi_model <- scvi$model$SCANVI$from_scvi_model(
    model,
    adata=adata_update,
    labels_key="subtype_standardized",
    unlabeled_category="Unknown",
)
scanvi_model

scanvi_model$train(accelerator = 'cpu', 
            devices = 'auto',
            max_epochs = 50L,
            train_size = 0.9)

# Saving consists of saving the model neural network weights, as well as parameters used to initialize the model.
model_dir <- "results/integrated_ref/scanvi"
scanvi_model$save(dir_path = model_dir, prefix = 'embryonic_adult_spinal_cord_motor_cortex_scanvi_', save_anndata = TRUE, overwrite=TRUE)
scanvi_model
```

```{r}
scanvi_dir <- "results/integrated_ref/scanvi/"
scanvi_model <- scvi$model$SCANVI$load(dir_path = scanvi_dir, 
                                       prefix = "embryonic_adult_spinal_cord_motor_cortex_scanvi_")

adata_update <- anndata::read_h5ad(file.path(scanvi_dir,"embryonic_adult_spinal_cord_motor_cortex_scanvi_umap_adata.h5ad"))

# Add the scanvi latent data to the adata object 
scanvi_latent <- scanvi_model$get_latent_representation(adata_update)
scanvi_mde <- scvi$model$utils$mde(data = scanvi_latent)
adata_update <- AnnData(
  X = adata_update$X,
  obs = adata_update$obs, 
  var = adata_update$var, 
  uns = adata_update$uns,
  obsm = list("_scvi_extra_categorical_covs" = adata_update$obsm$`_scvi_extra_categorical_covs`,
              "X_scVI" = adata_update$obsm[['X_scVI']], 
              "X_scANVI" = scanvi_latent, 
              "X_scANVI_MDE" = scanvi_mde), 
  layers = list("counts" = adata_update$layers['counts'],
                "scvi_normalized" = adata_update$layers['scvi_normalized'])
)
```

```{r}
sc$pp$neighbors(adata_update, use_rep='X_scANVI')
sc$tl$umap(adata_update, min_dist=0.3)

# anndata::write_h5ad(adata_update, filename = "results/integrated_ref/scanvi/embryonic_adult_spinal_cord_motor_cortex_scanvi_umap_adata.h5ad")

# pdf("figures/cell_refs/allen_yadav_rayon_integrated/subset/scanvi/scanvi_subtype_standardized_scanvi_UMAP.pdf", height = 15, width = 20)
sc$pl$umap(
    adata_update,
    color=c("subtype_standardized"),
    frameon=TRUE,
)
# dev.off()

# pdf("figures/cell_refs/allen_yadav_rayon_integrated/subset/scanvi/scanvi_dataset_donor_scanvi_UMAP.pdf", height = 10, width = 20)
sc$pl$umap(
    adata_update,
    color=c('dataset','donor'),
    ncols=2L,
    frameon=TRUE,
)
# dev.off()

# pdf("figures/cell_refs/allen_yadav_rayon_integrated/subset/scanvi/scanvi_motorneuron_scanvi_UMAP.pdf", height = 10, width = 12)
sc$pl$umap(
    adata_update,
    color=c('motorneuron'),
    frameon=TRUE,
)
# dev.off()
```

# Differential Expression

```{r}
# Found 13 SCT models. Recorrecting SCT counts using minimum median counts: 2842
merged_split_obj <- PrepSCTFindMarkers(merged_split_obj)
```

```{r}
Idents(merged_split_obj) <- merged_split_obj$subtype_standardized
marker_genes <- FindAllMarkers(merged_split_obj,
                               test.use = "negbinom", 
                               min.pct = 0.2,
                               return.thresh = 0.001,
                               min.cells.feature = 40)
```

# Figures

### Integrated Layers

```{r}
cell_type_table <- merged_split_obj@meta.data %>% 
  group_by(dataset, subtype_standardized) %>% 
  dplyr::count() %>% 
  pivot_wider(names_from = dataset, values_from = n) %>% 
  mutate(across(Allen_M1:yadav_2023, ~ifelse(is.na(.), 0,.)))

View(cell_type_table)
```

TO DO: - plot expression of the genes in the Yadav dorsal and ventral
clusters - plot SST cell-type expression - Add Allen institute donor age

```{r}
merged_split_obj
Idents(merged_split_obj) %>% head()
```

```{r}
colors_mn <- c("Other"="grey80", "Motoneurons" = "#641A80FF", pMN = "#B63679FF" ,MN = "#FECE91FF")

reds <- c("umap.unintegrated", "umap.harmony", "umap.rpca")
meta_cols <- c("motorneuron", "age_clean", "donor", "tissue","top_level_annotation", "top_level_standardized", "class_label_standardized", "subtype_standardized")
subdir <- c("subset")
to_plot <- crossing(reds, meta_cols)

purrr::map(1:nrow(to_plot), function(x){
  to_plot <- to_plot %>% 
    dplyr::slice(x)
  red <- to_plot[["reds"]]
  col <- to_plot[["meta_cols"]]

  outfile <- glue("figures/cell_refs/allen_yadav_rayon_integrated/{subdir}/embryonic_adult_spinal_cord_motor_cortex_{gsub('umap.','', red)}_integrated_{col}_UMAP.pdf")
  cc <- ifelse(grepl("motorneuron", col), "colors_mn", "colors_vector")
  dp <- DimPlot(merged_split_obj,
        group.by = col,
        reduction = red,
        label = TRUE,
        repel = TRUE) +
    scale_color_manual(values = get(cc)) +
    theme(legend.position = "bottom")
  
  pdf(outfile, height = 15, width = 15)
  print(dp)
  dev.off()
})
```


```{r}
table(toupper(blum_alpha_mn_degs$X1) %in% rownames(merged_split_obj))
table(toupper(blum_gamma_mn_degs$X1) %in% rownames(merged_split_obj))
```

```{r}
Idents(merged_split_obj) <- merged_split_obj$top_level_standardized
alpha <- intersect(toupper(blum_alpha_mn_degs$X1), rownames(merged_split_obj))
gamma <-  intersect(toupper(blum_gamma_mn_degs$X1), rownames(merged_split_obj))

merged_split_obj <- ScaleData(merged_split_obj, features = gamma)
# Single cell heatmap of feature expression
pdf("figures/cell_refs/allen_yadav_rayon_integrated/subset/gamma_motorneurons_marker_degs.pdf", height = 25, width = 15)
DoHeatmap(subset(merged_split_obj, downsample = 1000),
          assay = "SCT",
          features = gamma,
          size = 3)
dev.off()

pdf("figures/cell_refs/allen_yadav_rayon_integrated/subset/alpha_motorneurons_marker_degs.pdf", height = 25, width = 15)
merged_split_obj <- ScaleData(merged_split_obj, features = alpha)
DoHeatmap(subset(merged_split_obj, downsample = 1000), 
          assay = "SCT",
          features = alpha,
          size = 3)
dev.off()
```

```{r fig.height=15, fig.width=15,message=FALSE, warning=FALSE}
Idents(merged_split_obj) <- merged_split_obj$subtype_standardized

purrr::map(names(marker_genes_chun_list), function(x){
    FeaturePlot(merged_split_obj,
            features = marker_genes_chun_list[[x]],
            reduction = "umap.harmony",
            label.size = 2,
            label = TRUE,
            repel = TRUE) +
    theme(legend.position = "bottom") +
    patchwork::plot_annotation(title = x)
})
```






## Neuronal Subtypes

```{r fig.height=10, fig.width=10}
bc <- merged_split_obj@meta.data %>% 
  filter(grepl("neuron|motor", top_level_standardized)) %>% 
  pull(barcode)

neuron_subtypes <-  subset(merged_split_obj, cells = bc) 

DimPlot(neuron_subtypes,
        group.by = "top_level_standardized",
        reduction = "umap.harmony",
        # split.by = "dataset",
        label.size = 2,
        label = TRUE,
        repel = TRUE) +
    scale_color_manual(values = colors_vector) +
    theme(legend.position = "bottom")

DimPlot(neuron_subtypes,
        group.by = "top_level_standardized",
        reduction = "umap.harmony",
        split.by = "dataset",
        label.size = 2,
        label = TRUE,
        repel = TRUE) +
    scale_color_manual(values = colors_vector) +
    theme(legend.position = "bottom")
```

```{r fig.height=10, fig.width=10}
DimPlot(neuron_subtypes,
        group.by = "subtype_standardized",
        reduction = "umap.harmony",
        # split.by = "dataset",
        label.size = 2,
        label = TRUE,
        repel = TRUE) +
    scale_color_manual(values = colors_vector) +
    theme(legend.position = "bottom")

DimPlot(neuron_subtypes,
        group.by = "subtype_standardized",
        reduction = "umap.harmony",
        split.by = "dataset",
        label.size = 2,
        label = TRUE,
        repel = TRUE) +
    scale_color_manual(values = colors_vector) +
    theme(legend.position = "bottom")
```

```{r fig.height=7, fig.width=20}
Idents(neuron_subtypes) <- neuron_subtypes$donor
FeaturePlot(neuron_subtypes,
            features = "SST",
        # group.by = "top_level_standardized",
        reduction = "umap.harmony",
        split.by = "dataset",
        label.size = 2,
        label = TRUE,
        repel = FALSE) +
    theme(legend.position = "bottom")

FeaturePlot(neuron_subtypes,
            features = "SST",
        # group.by = "top_level_standardized",
        reduction = "umap.harmony",
        label.size = 2,
        label = TRUE,
        repel = FALSE) +
    theme(legend.position = "bottom")
```

## Motor Neurons and subsets


```{r}
Idents(merged_split_obj) <- merged_split_obj$subtype_standardized

bc <- merged_split_obj@meta.data %>% 
  filter(grepl("motor|pMN|astro|oligo|opc", subtype_standardized)) %>% 
  rownames(.)

subset_common_types <- subset(merged_split_obj, cells = bc)
Idents(subset_common_types) <- subset_common_types$subtype_standardized
table(Idents(subset_common_types))
```


```{r fig.height=10, fig.width=10}
# pdf("figures/cell_refs/embryonic_adult_spinal_cord_motor_cortex_integrated_harmony_subset_UMAP.pdf", height = 15, width = 11)
DimPlot(subset_common_types, 
        label = TRUE,
        repel = TRUE,
        group.by = 'celltypes',
        # split.by = 'dataset',
        reduction = 'umap.harmony') +
  scale_color_manual(values = colors_vector) +
  theme(legend.position = 'bottom')
# dev.off()

# pdf("figures/cell_refs/embryonic_adult_spinal_cord_motor_cortex_integrated_rpca_subset_UMAP.pdf", height = 15, width = 11)
DimPlot(subset_common_types,
        label = TRUE,
        repel = TRUE,
        group.by = 'celltypes',
        reduction = 'umap.rpca') +
  scale_color_manual(values = colors_vector) +
  theme(legend.position = 'bottom')
# dev.off()
```

```{r}
alpha <- blum_alpha_mn_degs %>% 
  arrange(cluster) %>% 
  mutate(gene_name = toupper(X1)) %>% 
  filter(gene_name %in% rownames(merged_split_obj@assays$SCT))


gamma <-  blum_gamma_mn_degs %>% 
  arrange(cluster) %>% 
  mutate(gene_name = toupper(X1)) %>% 
  filter(gene_name %in% rownames(merged_split_obj@assays$SCT))
```

```{r fig.height=15, fig.width=15, message=FALSE, warning =FALSE}
purrr::map(unique(alpha$cluster), function(clust){
  
  genes <- alpha %>% 
    filter(cluster == clust) %>% 
    pull(gene_name)
  n <- length(genes)
  genes <- genes[1:(min(n,10))]
  
  png(glue("figures/cell_refs/allen_yadav_rayon_integrated/subset/embryonic_adult_sc_mc_integrated_umap.harmony_blum_alpha_{clust}_UMAP.png"),
      height = 12, width = 12, res = 100, units = "in")
  print(FeaturePlot(subset_common_types,
            features = genes,
            reduction = "umap.harmony",
            label.size = 2,
            label = TRUE,
            repel = TRUE) +
    theme(legend.position = "bottom") +
    patchwork::plot_annotation(title = paste0("Cluster", clust, " alpha MN")))
  dev.off()
})


purrr::map(unique(gamma$cluster), function(clust){
  
  genes <- gamma %>% 
    filter(cluster == clust) %>% 
    pull(gene_name) 
  n <- length(genes)
  genes <- genes[1:(min(n,10))]
  
  png(glue("figures/cell_refs/allen_yadav_rayon_integrated/subset/embryonic_adult_sc_mc_integrated_umap.harmony_blum_gamma_{clust}_UMAP.png"),
      height = 12, width = 12, res = 100, units = "in")
  print(FeaturePlot(subset_common_types,
            features = genes,
            reduction = "umap.harmony",
            label.size = 2,
            label = TRUE,
            repel = TRUE) +
    theme(legend.position = "bottom") +
    patchwork::plot_annotation(title = paste0("Cluster", clust, " gamma MN")))
  dev.off()
})
```

## Marker Gene Expression

```{r}
sel_markers <- names(rayon_marker_gene_list) %>%  grep("pMN|MN|_V[0-9]|rayon_Oligodendrocyte_NA", ., value = T)
sel_markers
```

```{r}
Idents(merged_split_obj) <- merged_split_obj$subtype_standardized
Idents(subset_common_types) <- subset_common_types$subtype_standardized

mn_feature_plots <- purrr::map(c("umap.rpca","umap.harmony"), function(red){
  purrr::map(sel_markers, function(x){
      out1 <- glue("figures/cell_refs/allen_yadav_rayon_integrated/subset/embryonic_adult_sc_mc_integrated_{red}_{x}_UMAP.png")
      fp1 <- FeaturePlot(merged_split_obj,
                       label = TRUE, 
                       repel = TRUE,
                       label.size = 2,
                       features = rayon_marker_gene_list[[x]],
                       reduction = red) +
      patchwork::plot_annotation(title = x)
    
      out2 <- glue("figures/cell_refs/allen_yadav_rayon_integrated/subset/embryonic_adult_sc_mc_integrated_{red}_{x}_subset_UMAP.png")
      fp2 <- FeaturePlot(subset_common_types,
                features = rayon_marker_gene_list[[x]],
                label = TRUE,
                repel = TRUE,
                reduction = red) +
      patchwork::plot_annotation(title = x)
      
      
      png(out1, units = "in", res = 100, height = 20, width = 20)
      print(fp1)
      dev.off()
      
      png(out2, units = "in", res = 100, height = 20, width = 20)
      print(fp2)
      dev.off()
  })

})
```

# Session Info

```{r}
devtools::session_info()
```
