---
title: "iPSC Motor Neurons Time-series Data"
author: "Jenny L Smith"
date: "`R Sys.Date()`"
output: html_document
---

# Set-up 

```{r set-up, echo=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80),
                      tidy=TRUE,
                      fig.align='center',
                      fig.width = 10, fig.height = 10,
                      eval = TRUE)

options(stringsAsFactors = FALSE, max.print = 100)
options(Seurat.object.assay.version = 'v5')
table = function (..., useNA = 'ifany') base::table(..., useNA = useNA)
```

```{r message=FALSE,warning=FALSE}
library(dplyr)
library(tidyr)
library(magrittr)
library(stringr)
library(tibble)
library(glue)

library(ggplot2)
library(plotly)
library(RColorBrewer)

library(Seurat)
library(reticulate)
library(anndata)
```

```{r}
conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter","dplyr")
conflicted::conflict_prefer("saveRDS","base")
```

# Define Functions

```{r}
colors_vector <- c(RColorBrewer::brewer.pal(12, "Paired"), 
            "blue1", "darkslategray3", "burlywood3", "#984EA3",
            "seagreen1", "yellow2", "orchid", "darkblue", 
            "lightsalmon2","slateblue1","lightskyblue4",
            "azure2", "chartreuse1",  "lemonchiffon2",
            "deeppink", "darkslategray1", "green4", "navajowhite", 
            "brown4", "darkgoldenrod2", "deepskyblue1", "lightpink") %>% 
  c(., ggpubr::get_palette("jco", 5)) %>% 
  c("mediumpurple4","magenta", "peru") %>% 
  c(ggsci::pal_igv()(51))

length(colors_vector)

sample_id_colors <- c("black",colors_vector[1:18])
```

```{r}
make_3D_umap <- function(metadata, idents_column, extra_label_cols=NULL, selected_colors=NULL){
    "
    idents_column is for labeling. ex) 'RNA_snn_res.0.7 as well as the UMAP colnames, eg UMAP_1, UMAP_2'
    "
    # 3D UMAP
    plot.data <- metadata %>% 
      select(all_of(c(idents_column)))
    plot.data$label <- paste(rownames(plot.data))
    if(!is.null(extra_label_cols)){
      plot.data$label <- paste(plot.data$label, 
                               metadata[[extra_label_cols]],
                               sep = "\n")
    }
    
    if(is.null(selected_colors)){
      selected_colors <- c("lightseagreen", "gray50",  "darkgreen", "red4", "red", "turquoise4","black","yellow4","royalblue1","lightcyan3",
                 "peachpuff3","lightskyblue","gray20","orange2","royalblue4","yellow3","gray80","darkorchid1","lawngreen","plum2", 
                 "darkorange4","darkmagenta","mediumorchid1","lightslateblue","midnightblue", "tan2") #26
    }

    cnames <- colnames(plot.data)
    formulas <- sapply(1:3, function(x) as.formula(paste0("~", grep(x, cnames, value = TRUE))))
    fig <- plot_ly(data = plot.data, 
        x = formulas[[1]], y = formulas[[2]], z = formulas[[3]], 
        color = as.formula(paste0("~", idents_column[4])), 
        colors = selected_colors, 
        type = "scatter3d", 
        mode = "markers", 
        marker = list(size = 1, width=2), # controls size of points
        text=~label, # This is that extra column we made earlier for which we will use for cell ID
        hoverinfo="text") #When you visualize your plotly object, hovering your mouse pointer over a point shows cell names

    return(fig)
}
```

```{r}
custom_dimplot <- function(metadata, colorby, dimred_colnames, custom_colors,
                           lab_size = 6,
                           ny = 0, nx = 0,
                           nbins = 4,
                           custom_lab_fill = NULL,
                           label_samples = TRUE,
                           color_text_grp = FALSE,
                           add_text = TRUE, 
                           seed = 2023){
  
  # define x and y column names
  x_names <- paste0(dimred_colnames, "_1")
  y_names <- paste0(dimred_colnames, "_2")
  
  # function to split the bins factor into numeric values
  find_max_bin <- function(bins){
    tab <- table(bins)
    tab_max <- tab[tab== max(tab)]
    
    names(tab_max)  %>%
      gsub("\\((.+),(.+)\\]", "\\1,\\2", .) %>% 
      str_split(., ",", n = 2) %>% 
      unlist() %>% 
      as.numeric() %>% 
      mean()
  }

  # define labels positions
  labs_df <- metadata %>% 
    select(all_of(c(colorby, x_names, y_names))) %>%
    # summarise(x = mean(!! as.name(x_names)),
    #           y = mean(!! as.name(y_names)))
    mutate(x_bins = cut(!! as.name(x_names), breaks = nbins),
           y_bins = cut(!! as.name(y_names), breaks = nbins)) %>% 
    group_by(!! as.name(colorby)) %>% 
    summarize(x = find_max_bin(x_bins), 
              y = find_max_bin(y_bins)) %>% 
    ungroup()
  
   # define point sizes based on how many total cells in data
   ncells <- nrow(metadata)
   pt_size <- case_when(
     ncells < 5e4 ~ 1,
     ncells >= 5e4 & ncells < 1e5 ~ 0.5,
     ncells >= 1e5 ~ 0.25)
   
   # Make the scatterplot
   plot <- ggplot(metadata, 
         aes_string(x = x_names, y = y_names)) +
    geom_point(aes_string(color = colorby),
               size = pt_size, alpha = 0.4) +
    scale_color_manual(values = custom_colors) +
    guides(color = guide_legend(override.aes = list(alpha = 1, size = 2))) +
    theme_classic() +
    theme(legend.position = "top")
   
   # customize label positions
   y_val <- ny
   x_val <- nx
   if(is.null(custom_lab_fill)){
     fill_val <-  "white"
   }else{
     fill_val <- custom_lab_fill
   }
   
   if(color_text_grp){
     mapping <- aes_string(x = "x", y = "y", label = colorby, color = colorby)

   }else{
     mapping <- aes_string(x = "x", y = "y", label = colorby)
   }
  
   if( all(add_text & label_samples) ){
     plot <- plot + 
           ggrepel::geom_text_repel(data = labs_df, 
                             mapping = aes_string(x = "x", y = "y", label = colorby),
                             color = "black",
                             nudge_y = y_val,
                              nudge_x = x_val,
                             segment.colour = NA,
                             size = lab_size,
                             seed = seed)
   }else if ( label_samples ) {
      plot <- plot + 
         ggrepel::geom_label_repel(data = labs_df, 
                           mapping = mapping,
                           fill = fill_val,
                           nudge_y = y_val,
                            nudge_x = x_val,
                           segment.colour = NA,
                           label.size = 0.5,
                           size = lab_size,
                           show.legend = FALSE,
                           seed = seed)
   }
   
   return(plot)
}
```

```{r}
custom_heatmaps <- function(seurat_obj, genes,columns = NULL, annot_colors = NULL, celltype="ALL"){
  
  if(celltype == "ALL"){
    cell_ids <- colnames(seurat_obj)
  }else{
    cell_ids <- CellsByIdentities(seurat_obj, idents = celltype) %>%  unlist()
  }
  
  print(c("The assay is", DefaultAssay(seurat_obj)))
  genes <- intersect(genes,rownames(seurat_obj))
  seurat_obj <- ScaleData(seurat_obj, features = genes)
  
  print("Extracting expression data")
  norm_counts <- LayerData(seurat_obj[[DefaultAssay(seurat_obj)]], 
                           layer = 'scale.data', 
                           features = genes, 
                           cells = cell_ids)
  
  col_annots <- seurat_obj@meta.data %>% 
    select(-all_of("barcode")) %>% 
    rownames_to_column("barcode") %>% 
    filter(barcode %in% colnames(norm_counts)) %>% 
    as.data.frame() %>%
    set_rownames(.$barcode) %>%
    select(all_of(columns))
  
  # row_annots <- temp %>% 
  #   set_rownames(.$coords) %>% 
  #   select(gene_name, p_val, p_val_adj) 
  
  x <- nrow(norm_counts)
  fontsize_rows <- case_when(
    x < 50 ~ 10,
    x >= 50 & x < 100 ~ 8, 
    x >= 100 & x < 200 ~ 6, 
    x >= 200 ~ 4)


  len <- 299
  colors <- colorRampPalette(viridis::viridis(10))(len)
  colors_list <- lapply(columns, function(x){
          grps <- unique(pull(col_annots, all_of(x)))
          n <- length(grps)
          colors <- annot_colors[1:n] %>% 
            set_names(grps)
          return(colors)
    })
  names(colors_list) <- columns
  

  pheatmap::pheatmap(mat=as.matrix(norm_counts), 
                     scale = "none", 
                     color = colors,
                     fontsize_row = fontsize_rows,
                     # main = glue::glue("{celltype}: DA Peaks p<0.05 & FC >= {th}"),
                     annotation_col = col_annots,
                     annotation_colors = colors_list,
                     # annotation_row = row_annots[,-1],
                     labels_row = rownames(norm_counts),
                     clustering_method = 'ward.D2',
                     show_colnames = FALSE)
}
```


# Parallelization 

```{r}
library(future)
# check the current active plan
plan()
```

```{r eval=FALSE}
plan("multisession", workers = 16)
plan()
options(future.globals.maxSize = 60000 * 1024^2)
```

# Genomic References 

```{r}
changho_refs_file <- "references/changho_updated_marker_genes_MN_IN_annotation.xlsx"
marker_genes_chun <- purrr::map_dfr(openxlsx::getSheetNames(changho_refs_file)[2:4], function(x){
  openxlsx::read.xlsx(changho_refs_file, sheet = x)
}) %>% 
  janitor::clean_names() %>% 
  distinct() %>% 
  mutate(group=paste(author, tolower(neural_pop)) %>% 
           gsub("\\s","_",.)) %>% 
  mutate(specific_markers_for_each_type_of_neuron=gsub("\\s","", specific_markers_for_each_type_of_neuron))

# marker_genes_chun
marker_genes_chun_list <- purrr::map(unique(marker_genes_chun$group), function(x){
  marker_genes_chun %>%
    filter(group == x) %>%
    pull(specific_markers_for_each_type_of_neuron)
}) %>% 
  set_names(unique(marker_genes_chun$group))
marker_genes_chun_list[["pan_neuronal_markers"]] <- str_split(marker_genes_chun[1,"pan_neuronal_markers"], pattern = ", ", n=3)[[1]]

# marker_genes_chun_list
```

```{r}
# https://github.com/regan-hamel/h-iPSCs-MNI/blob/master/R%20workflow_iMNs_LThiry_2020.Rmd
marker_genes_iPSC <- list(MNPCs = c("CHAT", "ISL1", "ISL2", "MNX1", "OLIG2", "NEUROG2"),
                INs = c("PAX3", "VSX2", "GATA3", "SOX14", "SIM1", "LBX1", "TLX3"),
                NPCs = c("SOX1", "SOX2", "MKI67"),
                Glial = c("SOX9", "S100B", "GFAP"),
                Oligodendrocytes = c("PDGFRA", "GALC"),
                iPSCs = c("NANOG", "POU5F1"))
```

```{r}
rayon_marker_genes <- openxlsx::read.xlsx(xlsxFile = "resources/cell_refs/Rayon_2021/TableS2.xlsx") %>% 
  janitor::clean_names() %>% 
  mutate(id = paste("rayon", type, neural_pop, sep="_") %>% 
           gsub("\\s", "_", .)) %>% 
  mutate(description = case_when(
    grepl("FP", id) ~ gsub("FP","floor plate", id),
    grepl("RP", id) ~ gsub("RP","roof plate", id),
    grepl("dp", id) ~ gsub("dp[0-9]", "dorsal interneuron progenitors", id),
    grepl("p[0-2]", id) ~ gsub("p[0-2]", "intermediate interneuron progenitors", id),
    grepl("p3", id) ~ gsub("p3","ventral interneuron progenitors p3", id), 
    grepl("V[0-3]", id) ~ gsub("V[0-9].?", "ventral neurons", id),
    grepl("MN", id) ~ gsub(".?MN","motor neuron", id),
    grepl("dl[0-6]", id) ~ gsub("dl[0-9]", "dorsal Interneuron", id),
    grepl("LTMR|LMTR", id) ~ gsub("C?-{0,1}(LTMR|LMTR).{0,}", "mechanoreceptor", id),
    grepl("Proprioceptor|nociceptor", id) ~ gsub("Peripheral","PNS Peripheral",id),
    grepl("Mesoderm|Eryt|Blood|Hema|Myoblast|Oligo", id) ~ paste("rayon",str_split_fixed(id, "_", n=3)[,2]),
    TRUE ~ id
  ) %>% gsub("_"," ", .) %>% gsub("rayon ","", .))


# View(rayon_marker_genes)
# head(rayon_marker_genes)

rayon_marker_genes_long <- rayon_marker_genes %>% 
  separate(genes_map_step1, into = paste0("gene",1:5), sep=", ") %>% 
  separate(genes_map_step2, into = paste0("gene",5:20), sep = ", ") %>% 
  pivot_longer(cols = matches("^gene"),
               names_to = "colname",
               values_to = "genes") %>% 
  filter(!is.na(genes)) 

# rayon_marker_genes_long

rayon_marker_gene_list <-  purrr::map(unique(rayon_marker_genes_long$id), function(x){
  rayon_marker_genes_long %>% 
    filter(id == x) %>% 
    pull(genes) %>% 
    unique()
})
names(rayon_marker_gene_list) <- unique(rayon_marker_genes_long$id)
# rayon_marker_gene_list
```

```{r}
rayon_degs <- openxlsx::read.xlsx("resources/cell_refs/Yadav_2023/1-s2.0-S0896627323000314-mmc5.xlsx", sheet = 1, check.names = FALSE, sep.names = " ") %>% 
  pivot_longer(cols = everything(),
               names_to = "subtype_annotation", 
               values_to = "degs") %>% 
  mutate(names = gsub("-[0-9]{1,2}$","", subtype_annotation) %>% 
           gsub(".+\\s(.+)$", "\\1", .) %>% 
           gsub("\\/|-","_", .))

esc_celltypes <- unique(rayon_degs$names)
rayon_deg_gene_list <-  purrr::map(esc_celltypes, function(x){
  rayon_degs %>%
    filter(names == x) %>%
    pull(degs) %>%
    unique()
})
names(rayon_deg_gene_list) <- esc_celltypes
```

```{r}
mn_subtype_markers <- openxlsx::read.xlsx("references/11122023_Alpha_Gamma_MNs_consolidated marker list.xlsx", sheet = 2)
head(mn_subtype_markers)
# table(mn_subtype_markers$celltype)
```


# Read in the Seurat Object

```{r}
#https://github.com/satijalab/seurat/issues/3622
options(future.rng.onMisuse="ignore")
```

```{r}
ipsc_obj <- 
  readRDS("results/day0_day50/seurat/iPSC_motor_neuron_day0-day50_time-course_filtered_merged_soupX_normalized_B124_seurat_obj_v5.RDS")
   # readRDS("results/day0_day50/seurat/iPSC_motor_neuron_day0-day50_time-course_filtered_merged_soupX_normalized_dimred_anno_seurat_obj_v5.RDS")
  # readRDS("results/day0_day50/seurat/iPSC_motor_neuron_day0-day50_time-course_filtered_merged_soupX_normalized_seurat_obj_v5.RDS")

# Add metadata
ipsc_obj$time_category <- case_when(ipsc_obj$batch == "batch1" ~ "early phase", 
                                    ipsc_obj$batch == "batch2" ~ "intermediate phase",
                                    ipsc_obj$batch == "batch3" ~ "late phase",
                                    ipsc_obj$batch == "batch4" ~ "late phase")
ipsc_obj@meta.data$time_point_day <- gsub("iPSC_|rerun_","", ipsc_obj@meta.data$id) %>% 
  factor(., levels = unique(.))


ipsc_obj
table(ipsc_obj$time_category)
table(ipsc_obj$time_point_day)
```

## Add Celltype Annotations

```{r}
if(is.null(ipsc_obj@meta.data$subtype_standardized)){
  annots <- read.csv("results/day0_day50/scanvi/iPSC_motor_neuron_day0-day50_time-course_scanvi_predictions.csv", row.names = 1)
  annots <- annots[colnames(ipsc_obj), ]
  ipsc_obj <- AddMetaData(ipsc_obj, metadata = annots)
}

table(ipsc_obj$subtype_standardized)
```

## Update Data Dimension Reduction 

min.dist	
This controls how tightly the embedding is allowed compress points together. Larger values ensure embedded points are moreevenly distributed, while smaller values allow the algorithm to optimise more accurately with regard to local structure. Sensible values are in the range 0.001 to 0.5.

n.neighbors	
This determines the number of neighboring points used in local approximations of manifold structure. Larger values will result in more global structure being preserved at the loss of detailed local structure. In general this parameter should often be in the range 5 to 50.

```{r eval=FALSE}
# Add metadata and idents
ipsc_obj$id <- factor(ipsc_obj$orig.ident, levels=unique(ipsc_obj$orig.ident))
ipsc_obj$seurat_clusters <- ipsc_obj$SCT_snn_res.1
Idents(ipsc_obj) <- ipsc_obj$orig.ident

# ipsc_obj
class(ipsc_obj[["RNA"]]) #v5

# Update umap params
ipsc_obj <- ipsc_obj %>% 
          RunUMAP(reduction = "pca",
                dims = 1:30,
                # features = VariableFeatures(ipsc_obj),
                assay = 'SCT',
                n.components = 3,
                n.neighbors = 30,# default 30
                metric = "cosine",
                min.dist = 0.01, # default 0.3
                n.epochs = 500,
                return.model = TRUE,
                verbose = T)

ipsc_obj <- AddMetaData(ipsc_obj,as.data.frame(ipsc_obj@reductions$umap@cell.embeddings))
# saveRDS(ipsc_obj,"results/day0_day50/seurat/iPSC_motor_neuron_day0-day50_time-course_filtered_merged_soupX_normalized_dimred_ipsc_obj_v5.RDS")
```


### Review Clustering plots

```{r fig.height=7, fig.width=7}
DimPlot(ipsc_obj, 
        reduction = "umap",
        label = TRUE,
        label.size = 5,
        repel = TRUE,
        dims = c(1,2)) +
  scale_color_manual(values = colors_vector) +
  theme(legend.position = 'bottom')


DimPlot(ipsc_obj, 
        reduction = "umap",
        group.by = 'SCT_snn_res.1',
        label = TRUE,
        label.size = 5,
        repel = TRUE,
        dims = c(1,2)) +
  scale_color_manual(values = colors_vector) +
  theme(legend.position = 'bottom')
```

```{r fig.height=10}
filt_d1 <- DimPlot(ipsc_obj, 
        group.by = c("batch","id"),
        label = T, 
        repel = T,
        label.size = 4) + 
  scale_color_manual(values = colors_vector)

filt_s1 <- DimPlot(ipsc_obj, 
        group.by = c("id"),
        split.by = "batch",
        label = T, 
        repel = T,
        label.size = 4) + 
  scale_color_manual(values = colors_vector)

# pdf("figures/day0_day50/filtered_merged_results/batches_124/iPSC_MN_day0-day50_filtered_merged_soupX_normalized_B124_batch_id_UMAP.pdf", height = 10, width = 15)
filt_d1
# dev.off()

# pdf("figures/day0_day50/filtered_merged_results/batches_124/iPSC_MN_day0-day50_filtered_merged_soupX_normalized_B124_split_batch_id_UMAP.pdf", height = 10, width = 15)
filt_s1
# dev.off()
```

## Overview 

```{r}
table(ipsc_obj$time_category)
```

```{r fig.height=10}
time_cat_umap <- custom_dimplot(metadata = ipsc_obj@meta.data, 
               dimred_colnames = "umap", 
               colorby = "time_category",
               lab_size = 10,
               add_text = FALSE,
               # ny = c(1.5,0,0),
               # nx = c(-1,0,0),
               color_text_grp = FALSE,
               # custom_lab_fill = c("black","grey90","white"),
               custom_colors = ggpubr::get_palette("simpsons", 3)) +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=14), 
        legend.title = element_blank(),
        legend.text = element_text(size = 18)) +
  guides(color = guide_legend(ncol=3, override.aes = list(size = 6, alpha = 1)))


# pdf("figures/day0_day50/filtered_merged_results/batches_124/iPSC_day0-day50_time_category_blackText_UMAP.pdf", height = 10, width = 10)
time_cat_umap
# dev.off()
```


```{r fig.height=15}
cats <- unique(ipsc_obj@meta.data$time_point_day)
y_vals <-  rep(0,length(cats)) %>% 
  set_names(cats)
y_vals['day35'] <- 0.75
y_vals[c('day45','day50')] <- -2

# x_vals <-  rep(0,length(cats)) %>% 
#   set_names(cats)
# x_vals[c('day45','day50')] <- 1

time_pt_umap <- custom_dimplot(ipsc_obj@meta.data, 
               dimred_colnames = "umap", 
               colorby = "time_point_day",
               lab_size = 6,
               ny = y_vals,
               # nx = x_vals,
               nbins = 15,
               add_text = FALSE,
               color_text_grp=TRUE,
               custom_colors = tail(colors_vector, n = 15)) +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=14), 
        legend.title = element_blank(),
        legend.text = element_text(size = 18)) +
  scale_y_continuous(breaks = seq(-7,7, by = 1)) +
  scale_x_continuous(breaks = seq(-8,6, by = 1)) +
  guides(color = guide_legend(ncol=8, override.aes = list(size = 6, alpha = 1)))


pdf("figures/day0_day50/filtered_merged_results/batches_124/iPSC_day0-day50_time_point_day_textColor_v2_UMAP.pdf", height = 10, width = 10)
time_pt_umap
dev.off()
```

## Annotations 

```{r}
ct_colors <- viridis::turbo(n = 8) %>% 
  set_names(value = c("astrocyte","dorsal\ninterneuron 3","dorsal\ninterneuron 4",
                      "microglia","motor neuron","neuroglial\nprogenitor","","oligodendrocytes"))

par(mar = c(10,2,2,2))
barplot(rep(1,length(ct_colors)), las = 2, names.arg = names(ct_colors), col = ct_colors)
```

```{r}
ipsc_obj@meta.data %>% 
  select(subtype_standardized, subtype_labels) %>% 
  distinct()
```

```{r fig.height=10, fig.width=10}
set.seed(2023)

ipsc_obj@meta.data$subtype_labels <- factor(ipsc_obj@meta.data$subtype_labels,
                                            levels = unique(ipsc_obj@meta.data$subtype_labels))
cell_cats <- unique(ipsc_obj@meta.data$subtype_labels)
y_vals <-  rep(0,length(cell_cats)) %>% 
  set_names(cell_cats)
y_vals['neuroglial\nprogenitor'] <- 2
y_vals['astrocyte'] <- -0.5
y_vals['dorsal\ninterneuron 3'] <- -1
y_vals['dorsal\ninterneuron 4'] <- 1
# y_vals

x_vals <-  rep(0,length(cell_cats)) %>% 
  set_names(cell_cats)
x_vals['neuroglial\nprogenitor'] <- -1
x_vals['oligodendrocytes'] <- -1

umap_scanvi_plot <- custom_dimplot(metadata = ipsc_obj@meta.data, 
               dimred_colnames = "umap", 
               colorby = "subtype_labels",
               add_text = FALSE,
               ny = y_vals,
               nx = x_vals,
               lab_size = 5,
               color_text_grp = TRUE,
               custom_lab_fill = c(rep("black",4), rep("white",1), rep("black",1),"white"),
               custom_colors = ct_colors) +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=14), 
        legend.title = element_blank(),
        # legend.key = element_rect(fill=NA),
        legend.text = element_text(size = 16)) +
  guides(color = guide_legend(ncol=4,override.aes = list(size = 4, alpha = 1)))
  
# pdf("figures/day0_day50/filtered_merged_results/batches_124/iPSC_day0-day50_scanvi_celltype_annots_UMAP.pdf", height = 10, width = 10)
umap_scanvi_plot
# dev.off()
```


## Motorneuron Markers

```{r eval=FALSE}
custom_heatmaps(seurat_obj = ipsc_obj, 
                genes = marker_genes_chun_list$rayon_motor_neurons,
                columns = c("SCT_snn_res.1", "time_point_day","time_category"),
                annot_colors = colors_vector)

custom_heatmaps(seurat_obj = ipsc_obj, 
                genes = marker_genes_chun_list$yadav_motor_neurons,
                annot_colors = c("SCT_snn_res.1", "time_point_day","time_category"))
```

```{r}
# not what I really expected
ipsc_obj <- AddModuleScore(ipsc_obj,
                           seed = 11272023,
                           features =  marker_genes_chun_list$rayon_motor_neurons)

FeaturePlot(ipsc_obj, features = paste0("Cluster",1:11))
```

scale.data stores the pearson residuals - these are good for downstream clustering/PCA/UMAP/integration/transfer.

data stores the corrected counts - these should be used for visualization/DE.

https://github.com/satijalab/seurat/issues/2503

```{r fig.height=10}
marker_genes <- marker_genes_chun_list[c("rayon_motor_neurons", "yadav_motor_neurons")]

plots <- purrr::map(names(marker_genes), function(geneset){
  genes <- marker_genes[[geneset]]
  
  print(c("The assay is", DefaultAssay(ipsc_obj)))
  genes <- intersect(genes, rownames(ipsc_obj))
  ipsc_obj <- ScaleData(ipsc_obj, features = genes)
  
  print("Extracting expression data")
  norm_counts <- LayerData(ipsc_obj[[DefaultAssay(ipsc_obj)]], 
                           layer = 'scale.data', 
                           features = genes)
  
  gene_colsums <- colSums(norm_counts)
  ipsc_obj$mean_zscore <- gene_colsums
  pct98 <- ipsc_obj$mean_zscore %>% quantile(., probs = 0.98)
  ipsc_obj$mean_zscore[ipsc_obj$mean_zscore >= pct98] <- pct98
  
  plot <- ggplot(ipsc_obj@meta.data, aes(x = umap_1, y = umap_2)) +
    geom_point(aes(color = mean_zscore), size = 0.5) +
    scale_color_gradient(low = "navy", high = "red") +
    labs(title = glue("Marker Gene Expression from {gsub('_',' ',geneset)}")) +
    theme_classic() +
    theme(plot.title = element_text(size=24),
          axis.title = element_text(size=24),
          axis.text = element_text(size=24), 
          legend.position = "top",
          legend.title = element_text(size = 20),
          legend.text = element_text(size = 24)) +
    guides(color = guide_colorbar(barwidth = 15, barheight = 1.5))
})
names(plots) <- names(marker_genes)

# png("figures/day0_day50/CTMR/iPSC_day0-day50_rayon_motor_neurons_mean_zscore_UMAP.png", height = 10, width = 10, units = "in", res = 150)
pdf("figures/day0_day50/CTMR/iPSC_day0-day50_rayon_motor_neurons_mean_zscore_UMAP.pdf", height = 10, width = 10)
plots$rayon_motor_neurons
dev.off()

# png("figures/day0_day50/CTMR/iPSC_day0-day50_yadav_motor_neurons_mean_zscore_UMAP.png", height = 10, width = 10, units = "in", res = 150)
pdf("figures/day0_day50/CTMR/iPSC_day0-day50_yadav_motor_neurons_mean_zscore_UMAP.pdf", height = 10, width = 10)
plots$yadav_motor_neurons
dev.off()
```


# Session Info

```{r}
session_info()
```






